# 1. Neural Network Validation

Compare the pulsatility measurements of the 49 patients in the repeatability study using three methods: human annotator #1 (Reader 1), human annotator #2 (Reader 2), and the neural network (nn-UNet, or NN) trained on ~200 separate studies. See `DeepCMR/README.md` for details of the training and prediction workflows.

```{r setup, include=FALSE}
# Common setup
source("setup.R")
```

## Interobserver Analysis: Neural Network vs. Two Humans

Data for the interobserver (sometimes refered to as repeatability) analysis on the test data set is loaded from data files generated by Jupyter notebook `DeepCMR/analysis/2_nnUNet_test_analysis.ipynb`.

```{r load_repeatability_data}
# Calculated areas for each frame
df_areas <- read_csv(PROJ_ROOT %//% "DeepCMR/results/nnUNet/repeatability_Task618/df_areas.csv.gz", show_col_types = FALSE)

# Pairwise interobserver biases for each study
df_bias <- read_csv(PROJ_ROOT %//% "DeepCMR/results/nnUNet/repeatability_Task618/df_bias.csv.gz", show_col_types = FALSE)

# Misc and QC metrics, not really used in the interobserver analysis here
df_metrics <- read_csv(PROJ_ROOT %//% "DeepCMR/results/nnUNet/repeatability_Task618/df_metrics.csv.gz", show_col_types = FALSE)
```

From the raw areas, extract the minimum area, maximum area, and calculate the pulsatility for each of the 3 readers:

```{r compare_by_reader}
# Make sure there are no duplicate patients in this dataset (there shouldn't be)
patients <- df_areas %>%
  select(`PatientName`) %>%
  unique()
numValidation <- nrow(patients)

# Helper function:
# take the areas dataframe, filter by Reader 1 or Reader 2, then group by PatientName and apply the function, f (either min() or max()). The final call to `rename()` changes the name of the calculated column to the name of the function that was used
area_minmax <- function(reader, fn) {
  func_name <- substitute(fn)
  df_areas %>%
    filter(Reader == reader) %>%
    select(-Reader) %>%
    group_by(`PatientName`) %>%
    summarize(fn(Area)) %>%
    rename(!!func_name := "fn(Area)")
}

# Create separate dataframes for each reader
r1_minmax <- patients %>%
  inner_join(area_minmax(1, max), by = "PatientName") %>%
  inner_join(area_minmax(1, min), by = "PatientName") %>%
  mutate(pulsatility = 100 * (max - min) / min)
r2_minmax <- patients %>%
  inner_join(area_minmax(2, max), by = "PatientName") %>%
  inner_join(area_minmax(2, min), by = "PatientName") %>%
  mutate(pulsatility = 100 * (max - min) / min)
nn_minmax <- patients %>%
  inner_join(area_minmax("DL", max), by = "PatientName") %>%
  inner_join(area_minmax("DL", min), by = "PatientName") %>%
  mutate(pulsatility = 100 * (max - min) / min)

# create three data frames for making pair-wise comparisons between R1, R2, and NN
r1r2_compare <- left_join(r1_minmax, r2_minmax, by = "PatientName")
r1nn_compare <- left_join(r1_minmax, nn_minmax, by = "PatientName")
r2nn_compare <- left_join(r2_minmax, nn_minmax, by = "PatientName")
```

Make scatter plots comparing pulsatility, max area, and min area by reader.

```{r make_scatter_compare}
fig.01a_reader_scatter_compare <- {
  # create a scatter plot
  gg_scatter_compare.pul <-
    partial(gg_scatter_compare,
      data.x = "pulsatility.x",
      data.y = "pulsatility.y",
      lim = c(10, 80),
      ticks.breaks = seq(10, 80, 20),
      title = bquote("Pulsatility:" ~ .(label.x) ~ "vs" ~ .(label.y)),
      units = "%"
    )
  p1 <- r1r2_compare %>%
    gg_scatter_compare.pul(
      label.x = "R1",
      label.y = "R2",
      line.color = "blue"
    )
  p2 <- r1nn_compare %>%
    gg_scatter_compare.pul(
      label.x = "R1",
      label.y = "NN",
      line.color = "green"
    )
  p3 <- r2nn_compare %>%
    gg_scatter_compare.pul(
      label.x = "R2",
      label.y = "NN",
      line.color = "red"
    )

  # Max Areas
  gg_scatter_compare.max <-
    gg_scatter_compare %>%
    partial(
      data.x = "max.x",
      data.y = "max.y",
      lim = c(3, 12),
      ticks.breaks = seq(3, 12, 3),
      title = bquote("Max Area:" ~ .(label.x) ~ "vs" ~ .(label.y)),
      units = cm^2
    )
  p4 <- r1r2_compare %>%
    gg_scatter_compare.max(
      label.x = "R1",
      label.y = "R2",
      line.color = "blue"
    )
  p5 <- r1nn_compare %>%
    gg_scatter_compare.max(
      label.x = "R1",
      label.y = "NN",
      line.color = "green"
    )
  p6 <- r2nn_compare %>%
    gg_scatter_compare.max(
      label.x = "R2",
      label.y = "NN",
      line.color = "red"
    )

  # Min Areas
  gg_scatter_compare.min <-
    gg_scatter_compare %>%
    partial(
      data.x = "min.x",
      data.y = "min.y",
      lim = c(2, 9),
      ticks.breaks = seq(2, 9, 2),
      title = bquote("Min Area:" ~ .(label.x) ~ "vs" ~ .(label.y)),
      units = cm^2
    )
  p7 <- r1r2_compare %>%
    gg_scatter_compare.min(
      label.x = "R1",
      label.y = "R2",
      line.color = "blue"
    )
  p8 <- r1nn_compare %>%
    gg_scatter_compare.min(
      label.x = "R1",
      label.y = "NN",
      line.color = "green"
    )
  p9 <- r2nn_compare %>%
    gg_scatter_compare.min(
      label.x = "R2",
      label.y = "NN",
      line.color = "red"
    )

  # `patchwork` layout
  (p1 | p2 | p3) /
    (p4 | p5 | p6) /
    (p7 | p8 | p9)
}
fig.01a_reader_scatter_compare
```

Reader 1 and Reader 2 exhibit fairly good agreement in pulsatility measurements across the 49 patients in the repeatability sub-study. There is some disagreement between readings, although as we will see in the next section, this is likely amplified by how pulsatility is defined. The agreements between minimum and maximum areas is significantly better. There is overall agreement between the NN and Reader 1, and between the NN and Reader 2. Similar comparisons between Reader 1, Reader 2, and the NN for the maximum and minimum areas of the pulmonary artery cross section show satisfactory agreement across the methods.

```{r save_scatter_compare, eval=F}
# Save the figure to file.
ajtools::ggfigsave(fig.01a_reader_scatter_compare, name = "validation_reader_scatter_compare", aspect_ratio = 1, path = "figures")
```

Create a figure comparing just the minimum and maximum areas in the test data set, and calculate the pairwwise intraclass correlation coefficients (ICC).

```{r pairwise_minmax_area_compare, fig.asp=1/3, eval=F}
scatter_facet <- function(data, label.x, label.y) {
  data <- 
    data %>%
    pivot_longer(
      cols = c(min.x, max.x),
      values_to = "min_max.x",
      names_to = c("min_max", "x_y"),
      names_sep = "\\."
    ) %>%
    mutate(
      "min_max.y" = case_when(
        min_max == "min" ~ min.y,
        min_max == "max" ~ max.y
      )
    ) %>%
    select(-max.y, -min.y)
  
  stats <- 
    data %>%
    mutate(
      pos.x = 0.15 * min(min_max.x) + 0.85 * max(min_max.x)
    ) %>%
    group_by(min_max) %>%
    summarize(
      pos.x = mean(pos.x),
      pos.y = 0.7 * min(min_max.y) + 0.3 * max(min_max.y),
      cor = cor(min_max.x, min_max.y, method = "pearson")^2,
      icc = irr::icc(cbind(min_max.x, min_max.y))$value
    ) %>%
    mutate(
      label = "atop(plain(ICC) == " %++% scibeautify(icc, 3) %++% ", italic(r^2) == " %++% scibeautify(cor, 3) %++% ")",
      label_icc = "plain(ICC) == " %++% scibeautify(icc, 3),
      label_r2 = "italic(r^2) == " %++% scibeautify(cor, 3)
    )
  
  theme_text <- calc_element("text", theme_get())
  
  data %>%
  ggplot() +
  aes(x = min_max.x, y = min_max.y) +
  geom_smooth(
    formula = y ~ x,
    method = "lm",
    se = T,
    color = "black",
    size = 0.4
  ) +
  geom_point(size = 0.7) +
  labs(
    y = NULL, #ajtools::bquote_remask(label.y ~ (cm^2)),
    x = NULL,
  ) +
  ggtitle(ajtools::bquote_remask(label.x ~ (cm^2))) +
  geom_text(
    data = stats,
    mapping = aes(x = pos.x, y = pos.y, label = label_icc),
    parse = T,
    size = 0.35 * theme_text$size,
    lineheight = theme_text$lineheight,
    fontface = theme_text$face,
    family = theme_text$family,
    colour = theme_text$colour,
  ) +
  facet_grid(
    rows = "min_max",
    switch = "y",
    scales = "free_y",
    space = "free_y"
  ) +
  scale_x_continuous(
    limits = c(2,11.75),
    breaks = seq(2,12,2),
    expand = expansion()
  ) +
  scale_y_continuous(
    breaks = seq(2,12,2)
  )
}

pairwise_minmax_area_compare <- patchwork::wrap_plots(
  scatter_facet(r1r2_compare, "R1 vs. R2 Areas") + ylab("Minimum Area Maximum Area"),
  scatter_facet(r1nn_compare, "R1 vs. NN Areas"),
  scatter_facet(r2nn_compare, "R2 vs. NN Areas")
) &
theme(
  strip.background = element_blank(),
  strip.text = element_blank(),
  strip.placement = "outside",
  panel.spacing = unit(10, "pt")
)

ajtools::ggfigsave(pairwise_minmax_area_compare, path = PROJ_ROOT %//% "figures", width = 6.5, aspect_ratio = 3, overwrite = T, open = T)
```

Bland-Altman Plots.

```{r make_bland_altman}
fig.01_bland_altman <- {
  b1 <- gg_bland_altman(
    r1r2_compare$pulsatility.x,
    r1r2_compare$pulsatility.y,
    plotTitle = "Pulsatility (R2-R1)",
    ciFillColors = c("steelblue3", "gray85")
  )
  b2 <- gg_bland_altman(
    r1nn_compare$pulsatility.x,
    r1nn_compare$pulsatility.y,
    plotTitle = "Pulsatility (NN-R1)",
    ciFillColors = c("darkolivegreen3", "gray85")
  )
  b3 <- gg_bland_altman(
    r2nn_compare$pulsatility.x,
    r2nn_compare$pulsatility.y,
    plotTitle = "Pulsatility (NN-R2)",
    ciFillColors = c("indianred3", "gray85")
  )
  b4 <- gg_bland_altman(
    r1r2_compare$max.x,
    r1r2_compare$max.y,
    plotTitle = "Max Area (R2-R1)",
    ciFillColors = c("steelblue3", "gray85")
  )
  b5 <- gg_bland_altman(
    r1nn_compare$max.x,
    r1nn_compare$max.y,
    plotTitle = "Max Area (NN-R1)",
    ciFillColors = c("darkolivegreen3", "gray85")
  )
  b6 <- gg_bland_altman(
    r2nn_compare$max.x,
    r2nn_compare$max.y,
    plotTitle = "Max Area (NN-R2)",
    ciFillColors = c("indianred3", "gray85")
  )
  b7 <- gg_bland_altman(
    r1r2_compare$min.x,
    r1r2_compare$min.y,
    plotTitle = "Min Area (R2-R1)",
    ciFillColors = c("steelblue3", "gray85")
  )
  b8 <- gg_bland_altman(
    r1nn_compare$min.x,
    r1nn_compare$min.y,
    plotTitle = "Min Area (NN-R1)",
    ciFillColors = c("darkolivegreen3", "gray85")
  )
  b9 <- gg_bland_altman(
    r2nn_compare$min.x,
    r2nn_compare$min.y,
    plotTitle = "Min Area (NN-R2)",
    ciFillColors = c("indianred3", "gray85")
  )

  (b1 | b2 | b3) /
    (b4 | b5 | b6) /
    (b7 | b8 | b9) &
    theme(
      title = element_text(size = 9)
    )
}
show(fig.01_bland_altman)
```

```{r save_bland_altman_plots, eval=F}
# Save the figure to file.
ajtools::ggfigsave(fig.01_bland_altman, aspect_ratio = 1, path = "figures")
```

## Bias and absolute bias comparisons

Now compare the pairwise bias (difference) and pairwise absolute bias (absolute difference) in the derived pulsatilities among the three readers (two humans and neural network. Even though the minimum and maximum areas show ~95% ICC, we expect there to be more variability in the interobserver pulsatility, since it is a derived measure that will propagate the variability in the minimum and maximum areas. We are looking to show that the bias between a human reader and the neural network is statistically similar to the bias between the two humans.

```{r bias_comparisons}
# Gather all the data columns into one column
# + calculate absolute bias
# + rename for better labeling on plots
df_bias_stats <-
  df_bias %>%
  rename("R1 vs R2" = bias_R1_vs_R2, "R1 vs NN" = bias_R1_vs_DL, "R2 vs NN" = bias_R2_vs_DL) %>%
  gather(method, bias, -PatientName) %>%
  mutate(method = fct_relevel(method, "R1 vs R2", "R1 vs NN", "R2 vs NN"))

# ANOVA calculation
aov_tbl <-
  df_bias_stats %>%
  mutate(absbias = abs(bias))
p_aov_bias <- scibeautify(summary(aov(bias ~ method, data = aov_tbl))[[1]][1, "Pr(>F)"], sig_digits = 2)
p_aov_absbias <- scibeautify(summary(aov(absbias ~ method, data = aov_tbl))[[1]][1, "Pr(>F)"], sig_digits = 2)

# Create figure of bias distributions
fig.01b_bias_distributions <-
  df_bias_stats %>%
  {
    # Plot a combo of boxplot, points (represented by dashed lines; shape = 95), and a half-violin
    gg_box_violin_base <- function(data, x, y) {
      ggplot(data) +
        aes(fill = rev({{ x }}), x = {{ x }}, y = {{ y }}) +
        geom_boxplot(outlier.shape = NA, width = 0.15, alpha = 0.5, position = position_nudge(-0.245)) +
        geom_point(size = 2, shape = 124, alpha = 0.7, position = position_nudge(-0.075)) +
        geom_half_violin(trim = F, scale = "count", alpha = 0.5, width = 0.7, position = position_nudge(0)) +
        xlab("") +
        theme(
          plot.title = element_text(hjust = 0.5),
          legend.position = "none",
          panel.grid.major.y = element_blank(),
          panel.grid.major.x = element_line()
        ) +
        coord_flip() +
        scale_x_discrete(limits = rev)
    }

    p1 <- identity(.) %>%
      gg_box_violin_base(x = method, y = bias / 100) +
      labs(title = bquote("Bias Distribution" ~ (p == .(p_aov_bias)))) +
      ylab(bquote("Pulsatility bias" ~ (Delta)))

    p2 <- mutate(., abs_bias = abs(bias)) %>%
      gg_box_violin_base(x = method, y = abs_bias / 100) +
      labs(title = bquote("Absolute Bias Distribution" ~ (p == .(p_aov_absbias)))) +
      ylab(bquote("Abs Pulsatility bias" ~ ("|" * Delta * "|")))

    # plot patchwork
    (p1 | p2)
  } %>%
  show()

# Print a table of statistics
tbl.01c_bias_stats <-
  df_bias_stats %>%
  mutate(bias = bias / 100) %>%
  group_by(Method = method) %>%
  summarize(across(
    .cols = bias,
    .fns = list(
      mean_bias = ~ mean(.x),
      sd_bias = ~ sd(.x),
      p_bias = ~ t.test(.x, mu = 0, alternative = "two.sided")$p.value,
      mean_absbias = ~ mean(abs(.x)),
      sd_absbias = ~ sd(abs(.x))
      # p_absbias = ~ t.test(abs(.x), mu = 0, alternative = "greater")$p.value
    )
  ))

tbl.01c_bias_stats %>%
  scibeautify(2, .cols = starts_with("bias")) %>%
  ajtools::kbl.escape(
    col.names = c("Comparison", "Mean", "S.D.", "P-value", "Mean", "S.D.")
  ) %>%
  kableExtra::kable_classic_2() %>%
  kableExtra::add_header_above(c("", "Bias" = 3, "Absolute Bias" = 2))
```

Calculate intraclass correlation coefficient for max and min areas:

```{r}
# All areas
df_areas %>%
  pivot_wider(names_from = Reader, values_from = Area) %>%
  select(-PatientName, -Phase, -`DL`) %>%
  psych::ICC()

# Min areas
df_areas %>%
  group_by(PatientName, Reader) %>%
  summarize(min_area = min(Area)) %>%
  ungroup() %>%
  pivot_wider(names_from = Reader, names_prefix = "Reader_", values_from = min_area) %>%
  select(-PatientName) %>%
  psych::ICC()

# Max areas
df_areas %>%
  group_by(PatientName, Reader) %>%
  summarize(max_area = max(Area)) %>%
  ungroup() %>%
  pivot_wider(names_from = Reader, names_prefix = "Reader_", values_from = max_area) %>%
  select(-PatientName) %>%
  psych::ICC()

# Pulsatility
df_areas %>%
  group_by(PatientName, Reader) %>%
  summarize(pulsatility = log(max(Area)/min(Area))) %>%
  ungroup() %>%
  pivot_wider(names_from = Reader, names_prefix = "Reader_", values_from = pulsatility) %>%
  select(-PatientName) %>%
  psych::ICC()
```

Calculate p-values for absolute bias comparisons:

```{r}
absbias_r1r2 <- df_bias_stats %>%
  filter(method == "R1 vs R2") %>%
  arrange(PatientName) %>%
  pull(bias) %>%
  abs()
absbias_r1nn <- df_bias_stats %>%
  filter(method == "R1 vs NN") %>%
  arrange(PatientName) %>%
  pull(bias) %>%
  abs()
absbias_r2nn <- df_bias_stats %>%
  filter(method == "R2 vs NN") %>%
  arrange(PatientName) %>%
  pull(bias) %>%
  abs()

absbias_r1r2_vs_r1nn <- t.test(x = absbias_r1r2, y = absbias_r1nn, paired = T)$p.value
absbias_r1r2_vs_r2nn <- t.test(x = absbias_r1r2, y = absbias_r2nn, paired = T)$p.value
absbias_r1nn_vs_r2nn <- t.test(x = absbias_r1nn, y = absbias_r2nn, paired = T)$p.value

absbias_comparisons_pvalues <- tibble(
  absbias_r1r2_vs_r1nn,
  absbias_r1r2_vs_r2nn,
  absbias_r1nn_vs_r2nn
)
```

```{r save_bias_figs, eval=F}
ajtools::ggfigsave(fig.01b_bias_distributions, name = "validation_bias_distributions", aspect_ratio = 2, path = "figures")
```

## Training loss functions

Extract NN training information from the nnU-Net log files, and create a figure of loss functions:

```{r training_loss_function}
loss_fns <- tibble()
for (n in 0:4) {
  nnunet_log <- read_file(PROJ_ROOT %//%
    "DeepCMR/results/logs/Task618_UKBBPulmonaryArtery/nnUNet_train_14838.log/nnUNet_train_" %++%
    as.character(14834 + n) %++%
    "_success_fold" %++%
    as.character(n) %++%
    ".log")
  lf <- as.numeric(str_extract_all(nnunet_log, "(?<=train loss : )[-.\\d]+")[[1]])
  dice <- as.numeric(str_extract_all(nnunet_log, "(?<= Average global foreground Dice: \\[)[-.\\d]+")[[1]])
  loss_fns <- bind_rows(
    loss_fns,
    tibble(Epoch = 1:1000, LossFunction = lf, Dice = dice, Fold = rep(n, 1000))
  )
}

loss_fn_plot <-
  ggplot(loss_fns) +
  facet_grid(rows = vars(Fold)) +
  aes(x = Epoch, y = 1 + LossFunction) +
  geom_line() +
  scale_y_log10() +
  xlab("Epoch") +
  ylab("1 + Loss Function")
```

```{r, eval=F}
ajtools::ggfigsave(loss_fn_plot, path = "figures", width = 3, height = 3, open = T)
```

