# 2. Outlier Removal

Identification and removal of outliers (neural network prediction failures).

```{r setup, include=F}
# Common setup
source("setup.R")
```

Load data from the `df_dynamics`, which is generated by the Jupyter notebook `DeepCMR/analysis/3_nnUNet_predict_analysis.ipynb`.

```{r load_unfiltered_dynamics_data}
data_dir <- PROJ_ROOT %//% "DeepCMR/results/nnUNet/Task618_UKBBPulmonaryArtery/cmr_lvot_20212/"
df_dynamics <- read_csv(data_dir %//% "df_dynamics.csv.gz", show_col_types = F)
numPredicted <- nrow(df_dynamics)
```

## Define, characterize, and remove outliers

Data is loaded from data frames generated by the Jupyter notebooks. A total of **`r df_dynamics %>% count %>% as.integer %>% scibeautify`** studies were loaded.

```{r outliers_aaac, fig.asp=1/2}
outliers_aaac <- local({
  aaac <- df_dynamics$Total_abs_area_change_cm2 / 50
  upper <- mean(aaac, na.rm = T) + 3 * sd(aaac, na.rm = T)

  p <- df_dynamics %>%
    ggplot() +
    aes(x = Total_abs_area_change_cm2 / 50) +
    geom_histogram_aj() +
    scale_y_log10() +
    geom_vline(xintercept = upper, color = "orange", lty = "dashed") +
    ggtitle("Distrubition of Average absolute area change") +
    xlab(bquote("Average absolute area change" ~ (cm^2))) +
    ylab(bquote(log[10] ~ "Count"))

  p %>% show()

  df_dynamics %>%
    filter(Total_abs_area_change_cm2 / 50 > upper) %>%
    pull(PatientName)
})
```

There are `r outliers_aaac %>% length %>% scibeautify` out of `r df_dynamics %>% count %>% as.integer %>% scibeautify` studies that qualify as potential outliers in their value of the Average Absolute Area Change. The above graph shows the distribution (on a logarithmic scale) of the values for Average Absolute Area Change; the vertical dashed orange line corresponds to the outlier ($3\sigma$) cutoff.

In addition to Average Absolute Area Change, we'll also examine several other parameters, such as the Maximum Absolute Area Change (the maximum value for the absolute area change across all frames in a study). In summary, we will use the following as a list of heuristics to identify potential outliers:

1)  $z(\mathrm{Average~absolute~area~change}) > 3$
2)  $z(\mathrm{Maximum~absolute~area~change}) > 3$
3)  $z(\mathrm{Average~centroid~travel}) > 3$
4)  $z(\mathrm{Maximum~centroid~travel}) > 3$
5)  $\mathrm{Minimum~Area} = 0$
6)  $\mathrm{Ellipse~fit~success} = \mathrm{False}$

Applying these criteria to our data, we can tabulate the number of outliers:

```{r determine_outliers}
outliers <- list(
  "Average absolute area change $> 3\\sigma$" = local({
    aaac <- df_dynamics$Total_abs_area_change_cm2
    upper <- mean(aaac, na.rm = T) + 3 * sd(aaac, na.rm = T)
    df_dynamics %>%
      filter(Total_abs_area_change_cm2 > upper) %>%
      pull(PatientName)
  }),
  "Maximum absolute area change $> 3\\sigma$" = local({
    maac <- df_dynamics$Max_abs_area_change_cm2
    upper <- mean(maac, na.rm = T) + 3 * sd(maac, na.rm = T)
    df_dynamics %>%
      filter(Max_abs_area_change_cm2 > upper) %>%
      pull(PatientName)
  }),
  "Average centroid travel $> 3\\sigma$" = local({
    act <- df_dynamics$Centroid_total_travel_cm
    upper <- mean(act, na.rm = T) + 3 * sd(act, na.rm = T)
    df_dynamics %>%
      filter(Centroid_total_travel_cm > upper) %>%
      pull(PatientName)
  }),
  "Maximum centroid travel $> 3\\sigma$" = local({
    mct <- df_dynamics$Centroid_max_travel_cm
    upper <- mean(mct, na.rm = T) + 3 * sd(mct, na.rm = T)
    df_dynamics %>%
      filter(Centroid_max_travel_cm > upper) %>%
      pull(PatientName)
  }),
  "Minimum Area = 0" = {
    df_dynamics %>%
      filter(Min_area_cm2 == 0) %>%
      pull(PatientName)
  },
  "Failed ellipse fit" = {
    df_dynamics %>%
      filter(!Ellipse_fit_success) %>%
      pull(PatientName)
  }
)

# Rebuild as a data frame
outliers <- local({
  unique <- Reduce(union, outliers)
  df <- as_tibble(unique) %>% rename("Name" = 1)
  for (i in seq_along(outliers)) {
    heading <- names(outliers)[i]
    df %<>% mutate(!!heading := Name %in% outliers[[i]])
  }
  df
})
```

Save table of outliers to disk.

```{r save_outlier_candidates_table, eval=F}
write_csv(outliers, ajtools::file_datestamp("outlier_candidates.csv", path = data_dir))
```

Print a table summarizing outliers.

```{r make_table_of_outliers}
# Print table of counts
outliers %>%
  mutate("Total Unique Outliers" = T) %>%
  summarize(across(-Name, sum)) %>%
  transpose() %>%
  rename("Count" = 1) %>%
  scibeautify() %>%
  ajtools::kbl.escape(caption = "Number of candidate outliers based on six heuristics, selected to identify cases of potential image segmentation failure.") %>%
  kableExtra::kable_classic_2() %>%
  kableExtra::row_spec(row = 6, hline_after = T) %>%
  kableExtra::row_spec(row = 7, bold = T)
```

Examining a sample of the outliers, we can try to get a sense for what the failure modes of our network are. We first export our list of outliers from R, and then use python notebook `5_outlier_analysis.ipynb` to generate the following image previews. The image `outliers_001..020.png` is displayed below.

```{r outlier_preview}
knitr::include_graphics(
  PROJ_ROOT %//% "/DeepCMR/results/nnUNet/Task618_UKBBPulmonaryArtery/cmr_lvot_20212/outlier_previews/" %//% "outliers_001..020.png"
)
```

The image previews of each of the `r length(outliers[["Total Unique Outliers"]])` outlier candidates are manually inspected. Those studies that have ample evidence of acquisition failure (e.g. image corruption, incorrect view, slice out of plane of the LVOT) are labeled so we can get a sense of how often these occur.

```{r load_outlier_candidates_reviewed}
outliers <-
  read_csv(data_dir %//% "outlier_candidates_reviewed_09.08.21.csv", show_col_types = FALSE)

numOutliers <- nrow(outliers)

outliers %>%
  mutate(`Discard Reason` = as_factor(ifelse(is.na(`Unusable Reason`), "Other", `Unusable Reason`))) %>%
  group_by(`Discard Reason`) %>%
  summarize(Count = n()) %>%
  scibeautify(justify_mode = "d", .cols = Count) %>%
  ajtools::kbl.escape(align = c("l", "c")) %>%
  kableExtra::kable_classic()
```

A new dataframe with outliers removed is now saved.

```{r save_df_dynamics_minus_outliers, eval=F}
outliers <-
  read_csv(data_dir %//% "outlier_candidates.csv", show_col_types = FALSE)

numOutliers <- nrow(outliers)
df_dynamics_filtered <-
  df_dynamics %>%
  filter(!(PatientName %in% outliers$Name))

write_csv(df_dynamics_filtered, ajtools::file_datestamp("df_dynamics_minus_outliers.csv.gz", path = "data"))
```

Load the saved filtered data frame (outliers removed) from disk.

```{r load_df_dynamics_minus_outliers}
df_dynamics_filtered <-
  read_csv(PROJ_ROOT %//% "data/df_dynamics_minus_outliers_09.20.2021.csv.gz")

numPredicted_minusOutliers <- nrow(df_dynamics_filtered)
```

## Distribution of QC parameters

Here we look at the distributions of the pulsatilities, centroid travel, area changes, and ellipse parameters to see if there are any gross outliers that might represent failure of the neural network prediction.

```{r make_qc_distribution_plot}
fig.02a_qc_distributions_minus_outliers <-
  df_dynamics_filtered %>%
  {
    p1 <- ggplot(.) +
      aes(x = ageAtStudy) +
      geom_histogram_aj(binwidth = 1) +
      labs(
        title = "Age",
        x = "Age",
        y = "Count"
      )

    p2 <- ggplot(.) +
      aes(x = Max_area_cm2) +
      geom_histogram_aj(bins = 25) +
      xlim(c(0, 20)) +
      labs(
        title = "Maximum Area",
        x = bquote("Maximum Area" ~ (cm^2)),
        y = "Count"
      )

    p3 <- ggplot(.) +
      aes(x = Min_area_cm2) +
      geom_histogram_aj(bins = 25) +
      xlim(c(0, 20)) +
      labs(
        title = "Minimum Area",
        x = bquote("Minimum Area" ~ (cm^2)),
        y = "Count"
      )

    p4 <- ggplot(.) +
      aes(x = Max_area_frame) +
      geom_histogram_aj(bins = 50) +
      labs(
        title = "Max area frame",
        x = "Frame",
        y = "Count"
      )

    p5 <- ggplot(.) +
      aes(x = Min_area_frame) +
      geom_histogram_aj(bins = 50) +
      labs(
        title = "Min area frame",
        x = "Frame",
        y = "Count"
      )

    p6 <- ggplot(.) +
      aes(x = Total_abs_area_change_cm2) +
      geom_histogram_aj(bins = 20) +
      # scale_y_log10() +
      labs(
        title = "Cumulative Abs. Area Change",
        x = bquote("Area" ~ (cm^2)),
        y = bquote(Count)
      )

    p7 <- ggplot(.) +
      aes(x = Max_abs_area_change_cm2) +
      geom_histogram_aj(bins = 20) +
      # scale_y_log10() +
      labs(
        title = "Max (Step-wise) Abs. Area Change",
        x = bquote("Area" ~ (cm^2)),
        y = bquote(Count)
      )

    p8 <- ggplot(.) +
      aes(x = Centroid_total_travel_cm) +
      geom_histogram_aj(bins = 50) +
      # scale_y_log10() +
      labs(
        title = "Cumulative Centroid Travel ",
        x = bquote("Distance" ~ (cm)),
        y = bquote(Count)
      )

    p9 <- ggplot(.) +
      aes(x = Centroid_max_travel_cm) +
      geom_histogram_aj(bins = 50) +
      # scale_y_log10() +
      labs(
        title = "Max (Step-wise) Centroid Travel",
        x = bquote("Distance" ~ (cm)),
        y = bquote(Count)
      )

    p10 <- ggplot(.) +
      aes(x = FWHM_systole_frames) +
      geom_histogram_aj(bins = 25) +
      labs(
        title = "FWHM in Systole",
        x = "Frames",
        y = "Count"
      )

    # p1 /
    (p2 | p4) /
      (p3 | p5) /
      (p6 | p7) /
      (p8 | p9)
    # p10
  }
fig.02a_qc_distributions_minus_outliers
```

Save the QC distribution plot to disk.

```{r save_qc_distribution_plot, eval=F}
ajtools::ggfigsave(fig.02a_qc_distributions_minus_outliers)
```
