# 3. Merge CMR with UKB Data

We're now ready to link the CMR data for each patient with phenotypes from the UK Biobank phenotype dataset. Pulsatility data is loaded from the data file generated by the `02_outlier_removal.Rmd` notebook. Outliers have been removed from this data file. Patients are indexed by the `PatientName` field, which is the name attached to the DICOM files from the MRI images.

```{r setup, include=F}
# Common setup
source("setup.R")
```

Load CMR data minus outliers.

```{r load_data_minus_outliers}
df_dynamics <- read_csv(PROJ_ROOT %//% "data/df_dynamics_minus_outliers_09.20.2021.csv.gz", show_col_types = F)
```

We'll use a dictionary to convert "Patient Names" as coded in the CMR files to UKB participant identifiers. This dictionary was generated when bulk CMR data was first accessed from the UK Biobank servers. Of note, participants may have undergone MR imaging during Visit 2, Visit 3, or both (a fraction of patients are undergoing serial MR imaging.

```{r create_id_dictionary, eval=F}
# Helper/cleanup function:
# > standardize patient names so they match file names of MRI data
# > label studies that are duplicated (i.e. occur across multiple UKB_Instances)
dstr <- reticulate::import_from_path("strings", path = PROJ_ROOT %//% "DeepCMR/utils")
standardize_patient_name <- function(x) sapply(x, dstr$standardize_patient_name)
fix_instances <-
  Vectorize(function(this_id, this_inst, id_dict = id_dict) {
    matches <- which(id_dict$UKB_ID == this_id)
    num_matches <- length(matches)
    switch(as.character(num_matches),
      "1" = return(this_inst), # do nothing
      "2" = { # make sure inst = 2 and 3
        inst <- sort(id_dict$UKB_Instance[matches])
        stopifnot("Duplicate instances must be 2 and 3" = setequal(inst, c(2, 3)))
        return(toString(inst))
      },
      stop("ID: ", as.character(this_id), " has too many duplicates (found ", num_matches, ")")
    )
  }, c("this_id", "this_inst"))

# load local dictionary file & apply cleanup function
id_dict <-
  readxl::read_excel(PROJ_ROOT %//% "data/id-dictionaries/ukbcmr_20212_id_dictionary_022521.xlsx") %>%
  distinct(PatientName, .keep_all = T) %>%
  mutate(PatientName = standardize_patient_name(PatientName)) %>%
  mutate(
    .after = UKB_Instance,
    UKB_Instance_All = fix_instances(UKB_ID, UKB_Instance, .),
    isDuplicated = sapply(strsplit(UKB_Instance_All, ", "), length) > 1
  )

# Write the updated dictionary to disk.
write_rds(id_dict,
  ajtools::file_datestamp("ukbcmr_20212_id_dictionary_022521_with_duplicates.rds", path = "data/id-dictionaries"),
  "xz",
  compression = 9
)
```

Create a new dictionary with unique, first visits.

```{r make_id_dict_unique_first_visits, eval=F}
# Read the id dict (with duplicates) from disk
id_dict <- read_rds(PROJ_ROOT %//% "data/id-dictionaries" %//% "ukbcmr_20212_id_dictionary_022521_with_duplicates.rds")

# save unique, first visits to a new dictionary
id_dict_deduplicated <-
  id_dict %>%
  filter(!(isDuplicated & UKB_Instance != sapply(strsplit(UKB_Instance_All, ", "), function(s) sort(s)[1])))

# Save the new dictionary with unique, first visits to disk.
write_rds(id_dict_deduplicated,
  ajtools::file_datestamp("ukbcmr_20212_id_dictionary_022521_first_visits_only.rds", path = "data/id-dictionaries"),
  "xz",
  compression = 9
)
```

Read the de-duplicated dictionary from disk.

```{r load_id_dict_unique_first_visits}
id_dict_deduplicated <-
  read_rds(PROJ_ROOT %//% "data/id-dictionaries/ukbcmr_20212_id_dictionary_022521_first_visits_only.rds")
```

Create a table summarizing types of CMR visits.

```{r make_id_dict_summary_table}
# create a table summarizing visits
id_dict_deduplicated %>%
  inner_join(df_dynamics, by = "PatientName") %>%
  group_by("CMR Acquisition Visit" = paste0("Visit", ifelse(isDuplicated, "s ", " "), UKB_Instance_All, ifelse(isDuplicated, "", " only"))) %>%
  summarize("Number of participants" = n()) %>%
  bind_rows(summarize(., "CMR Acquisition Visit" = "Total Unique Participants", "Number of participants" = sum(`Number of participants`))) %>%
  scibeautify() %>%
  ajtools::kbl.escape() %>%
  kableExtra::kable_classic() %>%
  kableExtra::row_spec(., 4, bold = T)
```

The CMR data needs to be merged with the dictionary above in order to translate MRI-coded patient "names" into UKB participant IDs. Merge df_dynamics with de-duplicated id_dict:

```{r merge_id_dict_with_dynamics}
df_dynamics %<>%
  inner_join(id_dict_deduplicated, by = "PatientName")

numPredicted_minusOutliers_minusDuplicates <- nrow(df_dynamics)
```

Now merge df_dynamics with UKB data tables.

```{r merge_dynamics_with_ukb, eval=F}
df_merged <-
  df_dynamics %>%
  inner_join(read_rds(PROJ_ROOT %//% "data/ukb_pheno_subsets/CoreFields.rds")) %>%
  inner_join(read_rds(PROJ_ROOT %//% "data/ukb_pheno_subsets/Labs.rds")) %>%
  inner_join(read_rds(PROJ_ROOT %//% "data/ukb_pheno_subsets/CVOutcomes.rds")) %>%
  inner_join(read_rds(PROJ_ROOT %//% "data/ukb_pheno_subsets/DeathOutcomes.rds")) %>%
  inner_join(read_rds(PROJ_ROOT %//% "data/ukb_pheno_subsets/CMR.rds")) %>%
  inner_join(read_rds(PROJ_ROOT %//% "data/ukb_pheno_subsets/Hemodynamics.rds")) %>%
  inner_join(read_rds(PROJ_ROOT %//% "data/ukb_pheno_subsets/PhysicalActivity.rds")) %>%
  inner_join(read_rds(PROJ_ROOT %//% "data/ukb_pheno_subsets/Spirometry.rds"))

df_merged_extra <-
  df_dynamics %>%
  inner_join(read_rds(PROJ_ROOT %//% "data/ukb_pheno_subsets/PhysicalActivity.rds"))
# inner_join(read_rds(PROJ_ROOT %//% "data/ukb_pheno_subsets/Spirometry.rds"))
```

Save the merged df to disk for re-use in later notebooks. We'll use parquet files going forward.

```{r save_df_merged, eval=F}
ajtools::write_parquet_zstd(
  df_merged,
  ajtools::file_datestamp("ukb_cmr_pulsatility_merged", path = "data"),
  22
)

ajtools::write_parquet_zstd(
  df_merged_extra,
  ajtools::file_datestamp("ukb_cmr_pulsatility_merged_extra", path = "data"),
  22
)
```

Read the merged df from disk

```{r read_df_merged}
# df_merged <- arrow::read_parquet(PROJ_ROOT %//% "data" %//% "ukb_cmr_pulsatility_merged_2021_10_18.zst.parquet")
df_merged <- arrow::read_parquet(PROJ_ROOT %//% "data" %//% "ukb_cmr_pulsatility_merged.2021_11_19.zst.parquet")

numPredicted_minusOutliers_minusDuplicates_mergedWithUKB <- nrow(df_merged)
```
