# 5. Export Data for GWAS

In this notebook, we'll perform some QC-based exclusions on our CMR population to generate a GWAS cohort. We'll export all the required data for the GWAS (phenotype, covariates, principal components, etc.).

```{r setup, include=FALSE}
# Common setup
source("setup.R")
```

Load the merged data frame containing pulsatility, calculated genetic principal componenets, covariates, and UKB phenotype data.

```{r}
df_merged_pca <- arrow::read_parquet("data/ukb_cmr_pulsatility_and_flashPCA.2021_11_19.zst.parquet")
```

## Export GWAS Data

Define a function to export data to a phenotype and covariates file, as required for input into Bolt-LMM. Note that in the covariates file, participants are filtered-out based on reported/genetic sex mismatch and excessive kinship.

```{r generate_bolt_data_sets}
# Define function to write phenotypes and covariates to file in correct BOLT format
generate_bolt_data_sets <- function(data, file_prefix) {
  # PHENOTYPES
  # we only end up using the Pulsatility phenotype, but these others are there in case we want to look at these later
  data %>%
    transmute(
      FID = f.eid,
      IID = f.eid,
      Pulsatility = Pulsatility,
      BSA = ukbiobank::physio_bsa(., up_to_instance = UKB_Instance),
      EllipticalMaxArea = pi * (Minor_axis_at_max_cm)^2,
      EllipticalMinArea = pi * (Minor_axis_at_min_cm)^2,
      EllipticalMaxAreaIndex = EllipticalMaxArea / BSA,
      EllipticalMinAreaIndex = EllipticalMinArea / BSA,
      FWHM = FWHM_systole_frames,
      EccMax = Eccentricity_at_max
    ) %>%
    write_tsv(ajtools::file_datestamp("data/bolt-lmm/bolt_phenotypes." %++% file_prefix %++% ".tsv.gz"))

  # COVARIATES
  data %>%
    mutate(
      ageAtStudy = ukbiobank::age_at_instance(., instance_num = UKB_Instance),
      GenoArray = ifelse(f.22000.0.0.Genotype_measurement_batch > 0, "Axiom", "BiLEVE"),
      BMI = ukbiobank::physio_bmi(., up_to_instance = UKB_Instance),
      BSA = ukbiobank::physio_bsa(., up_to_instance = UKB_Instance),
      HR = ukbiobank::measurement_lookup_with_alt(.,
        measurement_col = f.22426.2.0.Average_heart_rate,
        measurement_col_alt = f.12673.2.0.Heart_rate_during_PWA,
        up_to_instance = UKB_Instance,
        combine_array = "mean"
      )
    ) %>%
    filter(
      f.31.0.0.Sex == f.22001.0.0.Genetic_sex,
      f.22021.0.0.Genetic_kinship_to_other_participants != "Ten or more third-degree relatives identified"
    ) %>%
    select(
      FID = f.eid,
      IID = f.eid,
      Age = ageAtStudy,
      Sex = f.31.0.0.Sex,
      GenoArray,
      HR,
      BMI,
      BSA,
      !!paste0("PC", 1:10)
    ) %>%
    write_tsv(ajtools::file_datestamp("data/bolt-lmm/bolt_covariates." %++% file_prefix %++% ".tsv.gz"))
}
```

Export general GWAS data (for our GWAS, we exclude patients with CHF, COPD, pHTN, and PE):

```{r, eval=F}
# Generate phenotype/covariate files for subjects:
# (minus PH, COPD, PE)
# (minus outliers identified in QC phase)
df_merged_pca %>%
  mutate(
    dx_CHF = ukbiobank::diagnosed_chf(., up_to_instance = UKB_Instance),
    dx_pHTN = ukbiobank::diagnosed_pHTN(., up_to_instance = UKB_Instance),
    dx_COPD = ukbiobank::diagnosed_copd(., up_to_instance = UKB_Instance),
    dx_PE = ukbiobank::diagnosed_pulm_embolism(., up_to_instance = UKB_Instance)
  ) %>%
  filter(
    !dx_CHF,
    !dx_pHTN,
    !dx_COPD,
    !dx_PE
  ) %>%
  generate_bolt_data_sets("flashPCA.minus_chf_phtn_copd_pe.with_BSA_and_HR")
```

Export data for sensitivity analysis (top 2.5% and bottom 2.5% percentiles in pulsatility removed):

```{r, eval=F}
# Generate phenotype/covariate files for subjects:
# (minus PH, COPD, PE)
# (minus outliers identified in QC phase)
df_merged_pca %>%
  mutate(
    dx_CHF = ukbiobank::diagnosed_chf(., up_to_instance = UKB_Instance),
    dx_pHTN = ukbiobank::diagnosed_pHTN(.),
    dx_COPD = ukbiobank::diagnosed_copd(.),
    dx_PE = ukbiobank::diagnosed_pulm_embolism(.)
  ) %>%
  filter(
    !dx_CHF,
    !dx_pHTN,
    !dx_COPD,
    !dx_PE,
    Pulsatility > quantile(Pulsatility, 0.025),
    Pulsatility < quantile(Pulsatility, 1 - 0.025)
  ) %>%
  generate_bolt_data_sets("minus_outliers_flashPCA_pHTN_COPD_PE_sensitivity")
```
