# 7. Analyze epidemiology

Perform epidemiologic analyses on PA pulsatility.

```{r setup, include=FALSE}
# Common setup
source("setup.R")
library(survival)

# custom function for defining range of boxplot whiskers -- span 95% of data
boxplot_range <- function(x) {
  r <- quantile(x, probs = c(0.025, 0.25, 0.5, 0.75, 0.975))
  names(r) <- c("ymin", "lower", "middle", "upper", "ymax")
  r
}
```

Load data:

```{r load_pulsatility}
df_merged <- arrow::read_parquet(PROJ_ROOT %//% "data/ukb_cmr_pulsatility_and_flashPCA.2021_11_19.zst.parquet")
```

## Define phenotypes

Create a new data frame with all of the phenotypes/metrics that will be of interest to us. This relies heavily on the `ukbiobank` package.

```{r define_phenotypes, eval=F}
df_pheno <-
  df_merged %>%
  transmute(
    f.eid,
    Pulsatility = Pulsatility,
    MRI_Instance = UKB_Instance,

    # demographics
    ethnicity = ukbiobank::ethnicity_self_reported(.),
    sex = f.31.0.0.Sex,
    age_at_mri = ukbiobank::age_at_instance_computed(., UKB_Instance),
    age_group_at_mri = ukbiobank::convert_to_age_group(age_at_mri),

    # death outcomes
    mri_to_death_yrs = ukbiobank::years_from_instance_to_death(., UKB_Instance),

    # comorbidities
    dx_chf = ukbiobank::diagnosed_chf(., up_to_instance = UKB_Instance),
    dx_pHTN = ukbiobank::diagnosed_pHTN(., up_to_instance = UKB_Instance),
    dx_copd = ukbiobank::diagnosed_copd(., up_to_instance = UKB_Instance),
    dx_htn = ukbiobank::diagnosed_htn(., up_to_instance = UKB_Instance),
    dx_hld = ukbiobank::diagnosed_hld(., up_to_instance = UKB_Instance),
    dx_diabetes = ukbiobank::diagnosed_diabetes(., up_to_instance = UKB_Instance),
    dx_pe = ukbiobank::diagnosed_pulm_embolism(., up_to_instance = UKB_Instance),

    # vitals
    sbp = ukbiobank::physio_systolicBP(.,
      up_to_instance = UKB_Instance,
      combine_instances = "last"
    ),
    dbp = ukbiobank::physio_diastolicBP(.,
      up_to_instance = UKB_Instance,
      combine_instances = "last"
    ),
    height = ukbiobank::physio_height_cm(.,
      up_to_instance = UKB_Instance,
      combine_instances = "last"
    ),
    weight = ukbiobank::physio_weight_kg(.,
      up_to_instance = UKB_Instance,
      combine_instances = "last"
    ),
    bmi = ukbiobank::physio_bmi(.,
      up_to_instance = UKB_Instance,
      combine_instances = "last"
    ),
    bsa = ukbiobank::physio_bsa(.,
      up_to_instance = UKB_Instance,
      combine_instances = "last"
    ),

    # crp
    crp = ukbiobank::biomarker_lookup(., f.30710.0.0.C_reactive_protein, up_to_instance = UKB_Instance)
  )
```

Write `df_pheno` to disk:

```{r save_df_pheno, eval=F}
write_rds(
  df_pheno,
  ajtools::file_datestamp(PROJ_ROOT %//% "data/ukb_cmr_pulsatility_Phenotypes.rds"),
  compress = "xz",
  compression = 9
)
```

Read `df_pheno`:

```{r read_df_pheno}
df_pheno <- arrow::read_parquet(PROJ_ROOT %//% "data/ukb_cmr_pulsatility_Phenotypes.2021_11_29.zst.parquet")
```

## Pulsatility by Category

### Pulsatility by Age

Plot pulsatility by age group as density plots:

```{r fig_pulsatility_by_age, fig.asp=1/2}
fig.06_pulsatility_by_age <-
  df_pheno %>%
  ggplot() +
  aes(x = Pulsatility, fill = age_group_at_mri) +
  geom_density(alpha = 0.7) +
  labs(
    # title="Pulsatility Density Plot by Age Group",
    x = "Pulsatility",
    y = "Density",
    fill = "Age Group"
  ) +
  xlim(0, 1) +
  theme(
    legend.justification = c("right", "top"),
    # legend.key.size = unit(4,"mm"),
    legend.position = c(0.99, 0.99),
    legend.box.margin = margin(),
    legend.key.height = unit(0.6, "lines"),
    legend.key.width = unit(1, "lines"),
    legend.title = element_text(margin = margin())
  ) +
  scale_fill_brewer()

fig.06_pulsatility_by_age
```

```{r, eval=F}
ajtools::ggfigsave(fig.06_pulsatility_by_age, path = "figures/", aspect_ratio = 2, overwrite = T)
```

Plot pulsatility by age group as box plots with AOV:

```{r fig_pulsatility_by_age_box, eval=F}
# AOV of age group
aov_agegroup <- TukeyHSD(aov(Pulsatility ~ age_group_at_mri, data = df_pheno))

# generate violin-boxplot of pulsatilities by age group
fig.06_pulsatility_by_age_box <-
  df_pheno %>%
  ggplot() +
  aes(y = Pulsatility, x = age_group_at_mri) +
  geom_violin(
    fill = "gray",
    alpha = 0.2,
    width = 0.85,
    adjust = 1.5,
    size = 0.4
  ) +
  stat_summary(
    fun.data = boxplot_range,
    geom = "boxplot",
    fill = "gray",
    alpha = 0.3,
    width = 0.25,
    color = "black",
    size = 0.4
  ) +
  ggsignif::geom_signif(
    comparisons = list(
      c(1, 2), c(2, 3), c(3, 4), c(4, 5)
    ),
    y_position = 0.9,
    step_increase = -0.03,
    extend_line = -0.02,
    map_signif_level = T,
    vjust = 0.5,
    test = "t.test",
    tip_length = 0.02,
    textsize = 0.38 * 7.5
  ) +
  scale_y_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1)) +
  xlab("Age Group (Years)")

# save figure to disk
ajtools::ggfigsave(fig.06_pulsatility_by_age_box, path = PROJ_ROOT %//% "figures", aspect_ratio = 2, overwrite = T)
```

### Pulsatility by Ethnicity

```{r, eval=F}
# box plot by ethnicity
fig.06_pulsatility_by_ethnicity_box <- print(
  df_pheno %>%
    ggplot() +
    aes(y = Pulsatility, x = ethnicity) +
    geom_boxplot(
      outlier.shape = NA,
      varwidth = F
    ) +
    ylim(0, 0.7) +
    xlab("Ethnicity")
)

# N and mean pulsatility by ethnicity
tbl.06_mean_pulsatility_by_ethnicity <- print(
  df_pheno %>%
    group_by(ethnicity) %>%
    summarize(
      N = n(),
      Pulsatility = mean(Pulsatility)
    )
)

# density plot of pulsatility by ethnicity
fig.06_pulsatility_by_ethnicity_density <- print(
  df_pheno %>%
    ggplot() +
    aes(x = age_at_mri, fill = ethnicity) +
    geom_density(alpha = 0.85) +
    labs(
      # title="Pulsatility Density Plot by Age Group",
      x = "Age",
      y = "Density",
      fill = "Age Group"
    ) +
    theme(
      legend.justification = c("right", "top"),
      legend.position = c(0.99, 0.99)
    )
)
```

### Pulsatility vs sex

```{r, eval=F}
df_pheno %>%
  ggplot() +
  aes(x = Pulsatility, fill = sex) +
  geom_density(alpha = 0.5) +
  labs(title = "Pulsatility Density Plot by Sex", x = "Pulsatility", y = "Density", fill = "Sex") +
  xlim(0, 1.2) +
  theme(
    legend.justification = c("right", "top"),
    legend.position = c(0.95, 0.95)
  )
```

### Pulsatility vs other

```{r, eval=F}
df_pheno %>%
  ggplot() +
  aes(x = Pulsatility, fill = dx_chf) +
  geom_histogram(alpha = 0.6) +
  scale_y_log10() +
  labs(title = "Pulsatility Density Plot by CHF Dx", x = "Pulsatility", y = "Count", fill = "") +
  xlim(0, 1.2)
```

Pulsatility vs HTN

```{r, eval=F}
df_pheno %>%
  ggplot() +
  aes(x = Pulsatility, fill = dx_htn) +
  geom_density(alpha = 0.5) +
  labs(title = "Pulsatility Density Plot by HTN", x = quote(Pulsatility), y = "Density", fill = "HTN Dx")
```

Pulsatility vs HLD

```{r, eval=F}
df_pheno %>%
  ggplot() +
  aes(x = Pulsatility, fill = dx_hld) +
  geom_density(alpha = 0.5) +
  labs(title = "Pulsatility Density Plot by HLD", x = quote(Pulsatility), y = "Density", fill = "HLD Dx")
```

Pulsatility vs COPD

```{r, eval=F}
df_pheno %>%
  ggplot() +
  aes(x = Pulsatility, fill = dx_copd) +
  geom_density(alpha = 0.5) +
  labs(title = "Pulsatility Density Plot by COPD", x = quote(Pulsatility), y = "Density", fill = "COPD Dx")
```

Pulsatility vs pHTN

```{r, eval=F}
df_pheno %>%
  ggplot() +
  aes(x = Pulsatility, fill = dx_pHTN) +
  geom_density(alpha = 0.5) +
  labs(title = "Pulsatility Density Plot by pHTN", x = quote(Pulsatility), y = "Density", fill = "pHTN Dx")
```

Pulsatility vs PE

```{r, eval=F}
df_pheno %>%
  ggplot() +
  aes(x = Pulsatility, fill = dx_pe) +
  geom_density(alpha = 0.5) +
  labs(title = "Pulsatility Density Plot by PE", x = quote(Pulsatility), y = "Density", fill = "PE Dx")
```

Pulsatility vs Diabetes

```{r, eval=F}
df_pheno %>%
  ggplot() +
  aes(x = Pulsatility, fill = dx_diabetes) +
  geom_density(alpha = 0.5) +
  labs(title = "Pulsatility Density Plot by Diabetes", x = quote(Pulsatility), y = "Density", fill = "Diabetes Dx")
```

Pulsatility vs Ethnicity

```{r, eval=F}
df_pheno %>%
  ggplot() +
  aes(x = Pulsatility, fill = ethnicity) +
  geom_density(alpha = 0.5) +
  labs(title = "Pulsatility Density Plot by White", x = quote(Pulsatility), y = "Density", fill = "White")
```

## Logistic Regressions with Comorbidities

```{r define_logistic_test_vars}
# Define a list of variables (comorbidities) that we would like to perform regression for
logistic_test_vars <- c(
  `Pulmonary Hypertension` = "dx_pHTN",
  CHF = "dx_chf",
  # Hyperlipidemia = "dx_hld",
  COPD = "dx_copd",
  Diabetes = "dx_diabetes",
  `Pulmonary Embolism` = "dx_pe",
  Hypertension = "dx_htn"
)

# perform logistic regression on each comorbidity, and store results in a list
glm_results <- lapply(logistic_test_vars, function(var) {
  df_pheno %>%
    mutate(
      is_female = sex == "Female",
      is_white = ethnicity == "White",
      Pulsatility = -100 * Pulsatility / 10,
    ) %>%
    glm(
      data = .,
      formula = bquote(.(as.symbol(var)) ~ age_at_mri + is_female + is_white + Pulsatility),
      family = binomial(link = "logit")
    )
})

# pass that list to ajtools::gg_glmforest to generate a nice forest plot/table
fig06.comorbidities_forest <-
  ajtools::gg_glmforest(
    glm_results,
    "Pulsatility",
    label.effect_size = "OR per 0.1 reduction in pulsatility",
    label.dependent_var = "Comorbidity",
    plot.rel_width = 0.35
  ) & theme(axis.text.x = element_text(size = rel(1)))

plot(fig06.comorbidities_forest)
```

```{r save_forest, eval=F}
# save forest plot to disk
ajtools::ggfigsave(fig06.comorbidities_forest, path = PROJ_ROOT %//% "figures", aspect_ratio = 3, open = F, overwrite = T)
```

## Survival analysis

Use some additional`ukbiobank::` functions to generate survival and time-to-event data to be used by the `survival` package:

```{r define_survival_data}
df_survival <-
  df_merged %>%
  transmute(
    # demographics
    is_white = ukbiobank::ethnicity_self_reported(.) == "White",
    sex = f.31.0.0.Sex,
    is_male = ifelse(sex == "Male", 1, 0),
    age_at_mri = ukbiobank::age_at_instance_computed(., UKB_Instance),

    # events
    mri_to_death_yrs = ukbiobank::years_from_instance_to_death(., UKB_Instance),
    mri_to_censor_yrs = ukbiobank::age_at_date(., "2021-04-01") - age_at_mri,
    status = ifelse(is.na(mri_to_death_yrs), 0, 1),
    time = ifelse(status == 0, mri_to_censor_yrs, mri_to_death_yrs),

    # pulsatility
    Pulsatility = Pulsatility,
    pulsatility_group = fct_relevel(case_when(
      Pulsatility >= 0.3 ~ "Pulsatility >= 30%",
      Pulsatility < 0.3 ~ "Pulsatility < 30%",
    ), c("Pulsatility >= 30%", "Pulsatility < 30%")),


    # Filters
    dx_chf = ukbiobank::diagnosed_chf(., up_to_instance = UKB_Instance),
    dx_copd = ukbiobank::diagnosed_copd(., up_to_instance = UKB_Instance),
    dx_diabetes = ukbiobank::diagnosed_diabetes(., up_to_instance = UKB_Instance),
    dx_phtn = ukbiobank::diagnosed_pHTN(., up_to_instance = UKB_Instance),

    # z-score
    pulsatility_z = -z_score(df_pheno$Pulsatility),
    age_z = z_score(df_pheno$age_at_mri),

    # Aliases
    `Male Sex` = is_male,
    `Age at MRI (years)` = age_at_mri,
    `Pulsatility decrease by 0.1` = -100 * Pulsatility / 10,
  )

# run the survival analysis
surv_fit <- survfit(Surv(time, status) ~ pulsatility_group, data = df_survival)
surv_diff <- survdiff(Surv(time, status) ~ pulsatility_group, data = df_survival)
logrank_p <- "Logrank * \":\" ~ italic(p) == " %++% scibeautify(1 - pchisq(surv_diff$chisq, sum(surv_diff$exp > 0) - 1), sig_digits = 2, output_format = "plotmath")
```

### Kaplan-Meier curves

Plot a Kaplan-Meier survival curve:

```{r make_survival_curve, eval=F}
fig.06_survival_curve <- ajtools::ggsurvplot_patchwork(
  surv_fit,
  axis.title.y.nudge = -40,
  risk.table = T,
  # cumevents = T,
  # cumcensor = T,
  conf.int = T,
  palette = rev(ggsci::pal_nejm()(2)),
  ylim = c(0.94, 1),
  legend.labs = c("Pulsatility \U2265 0.3", "Pulsatility < 0.3"),
  legend.title = "",
  legend = c(0.3, 0.6),
  xlim = c(-0.1, 6.5),
  xlab = "Time (years)",
  break.x.by = 1,
  censor = F
)

# add the logrank p-value to the plot
fig.06_survival_curve$patches$plots[[1]] <-
  fig.06_survival_curve$patches$plots[[1]] +
  patchwork::inset_element(
    grid::textGrob(str2expression(logrank_p), hjust = 0, vjust = 0.5, x = 0.325, y = 0.45),
    0, 0, 1, 1,
    align_to = "plot"
  )

# Save the plot to disk
ajtools::ggfigsave(fig.06_survival_curve, path = "figures/", open = T, aspect_ratio = 1.4, width = 5, overwrite = T)
```

### Cox proportional hazards

Calculate parameters of a Cox proportional hazards model:

```{r surv_table}
coxph_survival <- coxph(
  formula = Surv(time, status) ~ `Pulsatility decrease by 0.1` + `Age at MRI (years)` + `Male Sex`,
  data = df_survival
)

# Calculate parameters of a Cox proportional hazards model with only non-CHF and non-COPD patients:
# coxph_survival_healthy <- coxph(
#   formula = Surv(time, status) ~ `Pulsatility decrease by 0.1` + `Age at MRI (years)` + `Male Sex`,
#   data = df_survival %>% filter(!dx_copd & !dx_chf)
# )
```

Output the parameters from the coxph analysis:

```{r, fig.width=4, fig.height=1.5, eval=F}
coxph_survival_tableGrob <-
  coxph_survival %>%
  gtsummary::tbl_regression(exponentiate = T) %>%
  gtsummary::as_tibble() %>%
  rename_with(~ str_replace_all(.x, fixed("**"), "")) %>%
  rename("bolditalic(P) * bold(-value)" = "p-value") %>%
  mutate(`HR (95% CI)` = str_replace(`HR (95% CI)`, " to ", "–")) %>%
  gridExtra::tableGrob(
    d = .,
    rows = NULL,
    theme = gridExtra::ttheme_minimal(
      core = list(
        fg_params = list(
          hjust = matrix(c(0, 0.5, 0.5), nrow = nrow(.), ncol = 3, byrow = T),
          x = matrix(c(0, 0.5, 0.5), nrow = nrow(.), ncol = 3, byrow = T)
        )
      ),
      colhead = list(
        fg_params = list(
          hjust = c(0, 0.5, 0.5),
          x = c(0, 0.5, 0.5),
          parse = c(F, F, T)
        )
      )
    )
  ) %>%
  gtable::gtable_add_grob(
    grobs = grid::segmentsGrob(
      y1 = unit(0, "npc"),
      gp = grid::gpar(lwd = 2)
    ),
    t = 1, b = 1, l = 1, r = ncol(.)
  )
```

This is another Cox proportional hazards model where Pulsatility is stratified, and *adjusted survival curves* are displayed

```{r, eval=F, fig.width=5, fig.asp=1/1.4, eval=F}
cox <- coxph(
  formula = Surv(time, status) ~ pulsatility_group + age_at_mri + is_male + strata(pulsatility_group),
  data = df_survival
)
ajtools::ggsurvplot_patchwork(
  fit = survfit(cox),
  data = df_survival,
  axis.title.y.nudge = -20,
  conf.int = T,
  palette = rev(ggsci::pal_npg()(2)),
  ylim = c(0.92, 1),
  legend.labs = c("Pulsatility \U2265 25%", "Pulsatility < 25%"),
  legend.title = "",
  legend = c(0.2, 0.7),
  xlim = c(-0.1, 6.5),
  xlab = "Time (years)",
  break.x.by = 1,
  censor = F
)
```

## Violin-boxplot of comorbidities

```{r, eval=F}
bind_rows(
  df_pheno %>% filter(dx_copd) %>% mutate(dx = "COPD"),
  df_pheno %>% filter(dx_diabetes) %>% mutate(dx = "DM"),
  df_pheno %>% filter(dx_chf) %>% mutate(dx = "CHF"),
  df_pheno %>% filter(dx_pHTN) %>% mutate(dx = "Pulm\nHTN"),
  df_pheno %>% filter(!(dx_chf | dx_copd | dx_diabetes | dx_pHTN)) %>% mutate(dx = "Healthy"),
) %>%
  mutate(dx = as_factor(dx)) %>%
  select(dx, Pulsatility) %>%
  {
    ggplot(.) +
      aes(x = dx, y = Pulsatility) +
      geom_violin(
        fill = "gray",
        alpha = 0.2,
        width = 0.85,
        adjust = 1.5,
        size = 0.4
      ) +
      stat_summary(
        fun.data = boxplot_range,
        geom = "boxplot",
        fill = "gray",
        alpha = 0.3,
        width = 0.25,
        color = "black",
        size = 0.4
      ) +
      # geom_boxplot(
      #   outlier.shape = NA,
      #   na.rm = T,
      #   width = 0.25,
      #   color = "black",
      #   fill = "gray",
      #   alpha = 0.3,
      #   size = 0.4
      # ) +
      # ggforce::geom_sina(
      #   data =  ~filter(.x, dx != "No\ndisease"),
      #   maxwidth = 0.725,
      #   seed = pi/2
      # ) +
      scale_y_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1)) +
      xlab(NULL) +
      ggsignif::geom_signif(
        comparisons = rev(list(
          c(1, 5), c(2, 5), c(3, 5), c(4, 5)
        )),
        y_position = 0.72,
        extend_line = -0.025,
        map_signif_level = T,
        step_increase = 0.053,
        tip_length = 0.02,
        vjust = 0.6,
        textsize = 0.38 * 7.5,
        # parse = T
      ) +
      theme(
        axis.text = element_text(
          color = "black",
          size = calc_element("axis.title", theme_get())$size
        )
      )
  } %>%
  ajtools::ggfigsave(name = "fig.07_comorbidities_violin", path = PROJ_ROOT %//% "figures", width = 2.5, height = 1.5, format = c("pdf", "rds"), overwrite = T, open = T)
```


