# 9. eQTL colocalization

```{r setup}
# Common setup
source("setup.R")
library(locuscomparer)
```

## SETUP FOR COLOC

### Get the standard devation of our gwas trait (pulsatility), needed for coloc

```{r}
sdY_pulsatility <-
  arrow::read_parquet(PROJ_ROOT %//% "data/ukb_cmr_pulsatility_merged.2021_11_19.zst.parquet", col_select = "Pulsatility") %>%
  pull(Pulsatility) %>%
  sd()
```

### Load GWAS data

Load the GWAS data, including our candidate SNPs from FUMA, and our GWAS summary statistics. This is imported from another Rmd notebook:

```{r}
gwas_analysis <- ajtools::sourceRmd(PROJ_ROOT %//% "notebooks/08_analyze_gwas_results.Rmd")
fuma_cand_snps <- gwas_analysis$candidate_snps
fuma_risk_loci <- gwas_analysis$risk_loci
gwas_summary <- gwas_analysis$gwas_summary
remove(gwas_analysis)
```

### Helper functions

```{r}
# Find the `parquet` file that was download from GTEx/Google Cloud Platform (I converted them all to .zst.parquet, and added some annotations like hg19 coordinates and gene symbols).
gtex_tissue_chrN <- function(tissue = tissue_name, n) {
  if (n == 23) n <- "X"
  message("   > Loading GTEx data for chr", n, ", tissue ", tissue)
  arrow::read_parquet(
    PROJ_ROOT %//%
      "../../Archives.nosync/GTEx/GTEx_Analysis_v8_EUR_eQTL_all_associations/GTEx_Analysis_v8_EUR_eQTL_annotated" %//%
      as.character(tissue) %++% ".v8.EUR.allpairs.chr" %++% as.character(n) %++% "_annotated.zst.parquet",
    col_select = c(phenotype_id, external_gene_name, variant_id_b37, pval_nominal, slope, slope_se)
  )
}

# silent version of coloc::coloc.abf()
# -- suppress verbose messages from coloc.abf
runcoloc_silently <- function(...) purrr::quietly(coloc::coloc.abf)(...)$result

# silent, error-free version of coloc::runsusie()
# -- runsusie throws an error when it doesn't converge, which we can ignore and call it a negative result
# -- make sure runsusie is working as expected before using this error-free wrapper, though
runsusie_silently <- function(...) purrr::quietly(purrr::safely(coloc::runsusie))(...)$result$result

# custom patchwork legend for locuscompare
lc_legend <- {
  data.frame(
    expand.grid(x = 0:5, y = 0:1),
    fill = c(rep(NA, 6), "blue4", "skyblue", "darkgreen", "orange", "red", NA),
    color = c(rep(NA, 6), rep("black", 5), NA),
    label1 = c("", 0.2 * 1:4, rep("", 7)),
    label2 = c(rep("", 11), "r^2")
  ) %>%
    {
      ggplot(.) +
        geom_rect(aes(xmin = x, xmax = x + 1, ymin = y, ymax = y + 1), fill = .$fill, color = .$color) +
        geom_text(aes(x = x, y = y + 0.8), label = .$label1, vjust = 1, hjust = 0.5, size = 0.35 * calc_element("axis.title", theme_get())$size) +
        geom_text(aes(x = x + 0.2, y = y + 0.5), label = .$label2, parse = T, vjust = 0.5, hjust = 0, size = 0.35 * calc_element("axis.title", theme_get())$size) +
        scale_x_continuous(expand = expansion()) +
        scale_y_continuous(expand = expansion()) +
        theme(
          legend.position = "none",
          axis.line = element_blank(),
          axis.text = element_blank(),
          axis.title = element_blank(),
          axis.ticks = element_blank(),
          plot.margin = margin(),
          plot.background = element_rect(fill = "white", color = NA)
        )
    } %>%
    inset_element(0.02, 0.98 - 2 * 0.03, 0.02 + 18 * 0.03, 0.98, clip = F)
}
```

### Coloc parameters

```{r}
# GTEx tissues
tissues <- c(
  "Cells_Cultured_fibroblasts",
  "Heart_Left_Ventricle",
  "Heart_Atrial_Appendage",
  "Artery_Aorta",
  "Artery_Coronary",
  "Artery_Tibial",
  "Lung"
)

# minimum width for each risk locus
window_size <- 1e6 # 1 Mb

# where the plink1 files for 1000G data is stored, used when calculating LD r values
# --each file should be named chrN_phase3.bed/.bim/.fam, or chrN_phase3.tar.zst
genofiles_1000g_root <- PROJ_ROOT %//% "../../Archives.nosync/1000G/plink1_bed"

# where to look for/save LD tables for each locus
ld_table_lookup_root <- PROJ_ROOT %//% "analyses/LDtables"
```

## COLOC (single-variant)

Loop over all FUMA risk loci and perform coloc analysis

```{r, eval=F}
df_coloc_results <- tibble()

# output location
save_path_root <- PROJ_ROOT %//% "analyses/coloc_single_variant_v2"

for (tissue_name in tissues) {
  current_chr <- NULL
  gtex_tissue_chrN_file <- NULL

  ## ------------------------------------ loop over risk loci ------------------------------------
  for (risk_locus_n in fuma_risk_loci$GenomicLocus) {

    ## ---- get boundaries of risk locus ----
    risk_locus <- fuma_risk_loci %>% filter(GenomicLocus == risk_locus_n)
    chr <- risk_locus$chr
    len <- risk_locus$end - risk_locus$start
    # make the locus a minimum of window_size in width
    padding <- max(window_size - len, 0)
    min_pos <- risk_locus$start - padding / 2
    max_pos <- risk_locus$end + padding / 2
    message("> Processing risk locus ", risk_locus_n, " of ", nrow(fuma_risk_loci), " (", tissue_name, ", chr ", chr, ")")

    ## ---- load chr N gtex data ----
    if (!identical(current_chr, chr) || is.null(gtex_tissue_chrN_file)) {
      gtex_tissue_chrN_file <- gtex_tissue_chrN(tissue_name, chr)
      current_chr <- chr
    }

    ## ---- create subset of gtex and gwas data for this locus ----
    gwas_summary_risklocusN <-
      gwas_summary %>%
      filter(CHR == chr) %>%
      filter(BP >= min_pos, BP <= max_pos) %>%
      mutate(
        variant_id_b37 = paste(CHR, BP, ALLELE0_REF, ALLELE1_ALT, "b37", sep = "_")
      )

    gtex_tissue_risklocusN <-
      gtex_tissue_chrN_file %>%
      filter(variant_id_b37 %in% gwas_summary_risklocusN$variant_id_b37)

    ## ---- get LD info for these snps ----
    ld_cache_file <- ld_table_lookup_root %//% "LDtable_r_1000G_EUR_FUMAlocus" %++% formatC(risk_locus_n, width = 2, flag = "0") %++% "_" %++% "chr" %++% chr %++% ".zst.parquet"
    if (file.exists(ld_cache_file)) {
      ld <- arrow::read_parquet(ld_cache_file)
    } else {
      ld <- ajtools::ld_matrix_1000g_plink(
        query_snps = gwas_summary_risklocusN$SNP %>% str_replace("^([^:]+:\\d+)_.*", "chr\\1"),
        genofile = genofiles_1000g_root %//% "chr" %++% chr %++% "_phase3.tar.zst",
        pop = "EUR",
        mode = "r",
        sys.path = "/opt/homebrew/bin"
      ) %>%
        ajtools::write_parquet_zstd(ld_cache_file, 22)
    }
    ld_matrix <- ajtools::ld_list_to_matrix(ld, "SNP_A", "SNP_B", "R")
    remove(ld)

    ## ------------------------------------ loop over genes in the GTEx locus ------------------------------------
    cand_genes <- gtex_tissue_risklocusN$phenotype_id %>% unique()
    for (j in seq_along(cand_genes)) {
      search_gene <- cand_genes[j]

      ## ---- filter gtex_locus by gene ID ----
      gtex_locus_gene <- gtex_tissue_risklocusN %>%
        filter(phenotype_id == search_gene) %>%
        filter(!is.na(pval_nominal) & !is.na(slope) & !is.na(slope_se))

      ## ---- get common gene symbol ----
      gene_name <-
        if (is.na(gtex_locus_gene$external_gene_name[1]) || gtex_locus_gene$external_gene_name[1] == "") {
          gtex_locus_gene$phenotype_id[1] %>%
            as.character() %>%
            xfun::sans_ext()
        } else {
          gtex_locus_gene$external_gene_name[1] %>% as.character()
        }
      message("      > analyzing gene ", formatC(j, width = 3, flag = "0"), "/", formatC(length(cand_genes), width = 3, flag = "0"), ": ", gene_name)

      if (!any(gtex_locus_gene$pval_nominal < 0.05)) next

      ## ---- merge gtex and gwas loci by unique variant ID (allows for multi-allelic snps) ----
      locuscompare_data <- inner_join(gwas_summary_risklocusN, gtex_locus_gene, by = c("variant_id_b37"))
      remove(gtex_locus_gene)
      if (nrow(locuscompare_data) == 0) next

      ## ---- reformat data for coloc and locus compare ----
      locuscompare_data %<>%
        transmute(
          rsid = SNP,
          logp1 = -log10(P_BOLT_LMM_INF),
          logp2 = -log10(pval_nominal),
          chr = CHR,
          pos = BP,
          beta_gwas = BETA,
          varbeta_gwas = SE^2,
          beta_gtex = slope,
          varbeta_gtex = slope_se^2,
          variant_id_b37,
        ) %>%
        filter(
          !is.na(logp1) & !is.na(logp2) & !is.na(beta_gwas) & !is.na(varbeta_gwas) & !is.na(beta_gtex) & !is.na(varbeta_gtex)
        )

      ## ---- generate coloc data sets ----
      coloc_data_gwas <- list(
        beta = locuscompare_data$beta_gwas,
        varbeta = locuscompare_data$varbeta_gwas,
        snp = locuscompare_data$variant_id_b37,
        position = locuscompare_data$pos,
        type = "quant",
        sdY = sdY_pulsatility
      )
      coloc_data_gtex <- list(
        beta = locuscompare_data$beta_gtex,
        varbeta = locuscompare_data$varbeta_gtex,
        snp = locuscompare_data$variant_id_b37,
        position = locuscompare_data$pos,
        type = "quant",
        sdY = 1
      )

      ## ---- RUN COLOC ----
      coloc_res <- runcoloc_silently(dataset1 = coloc_data_gwas, dataset2 = coloc_data_gtex)
      coloc_H4_posterior <- coloc_res$summary["PP.H4.abf"]

      ## ---- store coloc results ----
      coloc_data <- tibble(
        fuma_locus_idx = risk_locus_n,
        gtex_tissue = tissue_name,
        gtex_gene = gene_name,
        gtex_ensembl_id = search_gene,
        locus_chr = chr,
        locus_start = min_pos,
        locus_end = max_pos,
        locus_min_gwasP = 10^(-(max(locuscompare_data$logp1))),
        locus_min_gtexP = 10^(-(max(locuscompare_data$logp2))),
        
        # store the posterior probabilities
        PP.H0 = coloc_res$summary["PP.H0.abf"],
        PP.H1 = coloc_res$summary["PP.H1.abf"],
        PP.H2 = coloc_res$summary["PP.H2.abf"],
        PP.H3 = coloc_res$summary["PP.H3.abf"],
        PP.H4 = coloc_res$summary["PP.H4.abf"],
        
        # store the snps with the highest snp-level posterior probability
        max_snp_PPH4_rsid = with(coloc_res$results, {
          locuscompare_data %>%
            filter(variant_id_b37 == snp[which.max(SNP.PP.H4)]) %>%
            pull(rsid)
        }),
        max_snp_PPH4 = with(coloc_res$results, SNP.PP.H4[which.max(SNP.PP.H4)]),
        
        # store the correlation between z-scores for the phenotype and the expression data
        zscore_cor = with(coloc_res$results, cor(z.df1, z.df2)),
        zscore_cor_top5pct = with(coloc_res$results %>% filter(SNP.PP.H4 > quantile(SNP.PP.H4, .95)), cor(z.df1, z.df2)),
      )

      # save results to dataframe
      df_coloc_results %<>% bind_rows(coloc_data)

      ## ------------------------------------ plot and save data if PP > 0.8 ------------------------------------
      if (coloc_H4_posterior > 0.8) {

        ## ---- save data ----
        save_path <-
          save_path_root %//%
          str_replace_all(tissue_name, "_", "") %++% "_" %++%
          "chr" %++% chr %++% "_" %++%
          "FumaRiskLocus" %++% formatC(risk_locus_n, width = 2, flag = "0")
        save_name <-
          str_replace_all(tissue_name, "_", "") %++% "_chr" %++% chr %++% "_" %++% gene_name
        dir.create(save_path, showWarnings = F, recursive = T)
        write_rds(coloc_res, save_path %//% save_name %++% ".coloc.rds", compress = "xz", compression = 9)

        ## ---- make sensitivity plots ----
        pdf(
          file = save_path %//% save_name %++% ".coloc.sensitivity.pdf",
          width = 6.5, height = 6.5,
          title = "coloc Sensitivity Analysis"
        )
        suppressMessages(coloc::sensitivity(coloc_res, rule = "H4 > 0.8"))
        dev.off()

        # identify the snp with the max PPH4
        max_row <- which.max(coloc_res$results$SNP.PP.H4)
        snp_max_PPH4 <- coloc_res$results$snp[max_row]
        snp_max_PPH4 <- locuscompare_data %>%
          filter(variant_id_b37 == snp_max_PPH4) %>%
          pull(rsid)

        ## ---- make locuscompare plot with each cand SNP ----
        message("         > saving plots")
        cand_snps_risklocusN <- fuma_cand_snps %>%
          filter(GenomicLocus == risk_locus_n) %>%
          pull(rsID)
        pb <- txtProgressBar(style = 3, min = 0, max = length(cand_snps_risklocusN))
        for (i in seq_along(cand_snps_risklocusN)) {
          snp_rsID <- cand_snps_risklocusN[i]

          setTxtProgressBar(pb, i)

          is_lead <- any(str_split_fixed(risk_locus$IndSigSNPs, ";", Inf)[1, ] == snp_rsID)
          is_maxPPH4 <- (snp_rsID == snp_max_PPH4)
          targetSNP_in_locus <- any(locuscompare_data$rsid == snp_rsID)

          if (targetSNP_in_locus) {
            save_name <-
              str_replace_all(tissue_name, "_", "") %++% "_chr" %++% chr %++% "_" %++% gene_name %++% "_" %++%
              str_replace_all(snp_rsID, ":", "_") %++% (if (is_lead) "_lead" else "") %++% (if (is_maxPPH4) "_maxPPH4" else "")

            r <- locuscomparer::make_combined_plot(
              locuscompare_data %>% select(rsid, logp1, logp2, chr, pos),
              title1 = "Pulsatility GWAS:",
              title2 = paste0("eQTL ", gene_name, " / ", str_replace_all(tissue_name, "_", " "), ":"),
              ld = tibble(SNP_A = snp_rsID, SNP_B = colnames(ld_matrix), R2 = ld_matrix[match(snp_rsID, colnames(ld_matrix)), ]^2),
              chr = chr,
              snp = snp_rsID,
              combine = FALSE,
              legend = FALSE
            )
            lc_patchwork <- (r$locuscompare + lc_legend) | (r$locuszoom1 / r$locuszoom2) & theme(axis.text = element_text(color = "black"))
            suppressMessages(invisible(ajtools::ggfigsave(
              lc_patchwork,
              name = save_name,
              path = save_path,
              format = c("png", "rds"),
              overwrite = T,
              width = 6.5,
              height = 6.5,
              dpi = 300
            )))
            remove(r, lc_patchwork)
          }
        }
        close(pb)
      }
    }
  }
}

# post-processing; attach some info from the FUMA file about the genomic risk loci & save to disk
df_coloc_results %<>%
  left_join(
    fuma_risk_loci %>%
      select(
        "idx" = GenomicLocus,
        start, end, nSNPs, LeadSNPs
      ) %>%
      rename_with(~ paste0("fuma_locus_", .x)),
    by = "fuma_locus_idx"
  ) %>%
  relocate(starts_with("fuma"), .after = "fuma_locus_idx") %>%
  write_tsv(ajtools::file_datestamp(basename(save_path_root) %++% "_results.txt.gz", path = save_path_root))
```

### Load coloc (single variant) results

```{r}
df_coloc_results <- vroom(PROJ_ROOT %//% "analyses/coloc_single_variant/coloc_single_variant_v2_results.2022_03_24a.txt.gz", show_col_types = F)
```

## COLOC-SuSiE

Loop over all FUMA risk loci and perform coloc with SuSiE. This is a variation on the above that allows for multiple causal variants. A credible set of SNPs is identified for the GWAS and eQTL traits, and the posterior probability of H4 for each pair of CSs is calculated.

```{r, eval=F}
df_coloc.susie_results <- tibble()

# where to save results and plots
save_path_root <- PROJ_ROOT %//% "analyses/coloc_SuSiE_50"

# a lower coverage allows us to detect weaker signals; USE CAUTION
susie_coverage <- 0.95

# ## for debugging
# tissue_name <- "Lung"
# risk_locus_n <- 10
# search_gene <- gtex_tissue_chrN(tissue_name, 12) %>% filter(external_gene_name == "PTHLH") %>% pull(phenotype_id) %>% `[`(1)

for (tissue_name in tissues) {
  current_chr <- NULL
  gtex_tissue_chrN_file <- NULL

  ## ------------------------------------ loop over risk loci ------------------------------------
  for (risk_locus_n in fuma_risk_loci$GenomicLocus) {
    risk_locus <- fuma_risk_loci %>% filter(GenomicLocus == risk_locus_n)
    chr <- risk_locus$chr
    message("> Processing risk locus ", risk_locus_n, " of ", nrow(fuma_risk_loci), " (", tissue_name, ", chr ", chr, ")")

    ## ---- get boundaries of risk locus ----
    # make the locus a minimum of window_size in width
    len <- risk_locus$end - risk_locus$start
    padding <- max(window_size - len, 0)
    min_pos <- risk_locus$start - floor(padding / 2)
    max_pos <- risk_locus$end + floor(padding / 2)

    ## ---- load annotated, tissue-specific, gtex data for chromosome N ----
    if (!identical(current_chr, chr) || is.null(gtex_tissue_chrN_file)) {
      gtex_tissue_chrN_file <- gtex_tissue_chrN(tissue_name, chr)
      current_chr <- chr
    }

    ## ---- create subsets of gwas and gtex data for this locus ----
    gwas_summary_risklocusN <-
      gwas_summary %>%
      filter(CHR == chr) %>%
      filter(BP >= min_pos, BP <= max_pos) %>%
      # create a unique variant ID in REF_ALT format (will use this to align with the gtex snps, which are labeled in this format)
      mutate(
        variant_id_b37 = paste(CHR, BP, ALLELE0_REF, ALLELE1_ALT, "b37", sep = "_")
      ) %>%
      filter(!is.na(P_BOLT_LMM_INF) & !is.na(BETA) & !is.na(SE))

    gtex_tissue_risklocusN <-
      gtex_tissue_chrN_file %>%
      filter(variant_id_b37 %in% gwas_summary_risklocusN$variant_id_b37) %>%
      filter(!is.na(pval_nominal) & !is.na(slope) & !is.na(slope_se))

    ## ---- get signed LD info for these snps ----
    ld_cache_file <- ld_table_lookup_root %//% "LDtable_r_1000G_EUR_FUMAlocus" %++% formatC(risk_locus_n, width = 2, flag = "0") %++% "_" %++% "chr" %++% chr %++% ".zst.parquet"
    if (file.exists(ld_cache_file)) {
      ld <- arrow::read_parquet(ld_cache_file)
    } else {
      ld <- ajtools::ld_matrix_1000g_plink(
        query_snps = gwas_summary_risklocusN$SNP %>% str_replace("^([^:]+:\\d+)_.*", "chr\\1"),
        genofile = genofiles_1000g_root %//% "chr" %++% chr %++% "_phase3.tar.zst",
        pop = "EUR",
        mode = "r",
        sys.path = "/opt/homebrew/bin"
      ) %>%
        ajtools::write_parquet_zstd(ld_cache_file, 22)
    }
    ld_matrix <- ajtools::ld_list_to_matrix(ld, "SNP_A", "SNP_B", "R")
    ld_not_missing <- apply(ld_matrix, 1, function(x) all(!is.na(x)))
    ld_matrix <- ld_matrix[ld_not_missing, ld_not_missing]
    remove(ld, ld_not_missing)

    ## ------------------------------------ loop over genes in the GTEx locus ------------------------------------
    cand_genes <- gtex_tissue_risklocusN$phenotype_id %>% unique()
    for (j in seq_along(cand_genes)) {
      search_gene <- cand_genes[j]

      ## ---- filter gtex_locus by gene ID ----
      gtex_locus_gene <-
        gtex_tissue_risklocusN %>%
        filter(phenotype_id == search_gene)

      ## ---- get gene symbol ----
      gene_name <-
        if (is.na(gtex_locus_gene$external_gene_name[1]) || gtex_locus_gene$external_gene_name[1] == "") {
          gtex_locus_gene$phenotype_id[1] %>%
            as.character() %>%
            xfun::sans_ext()
        } else {
          gtex_locus_gene$external_gene_name[1] %>% as.character()
        }
      message("      > analyzing gene ", formatC(j, width = 3, flag = "0"), "/", formatC(length(cand_genes), width = 3, flag = "0"), ": ", gene_name)

      if (!any(gtex_locus_gene$pval_nominal < 0.05)) next

      ## ---- merge gtex and gwas loci by unique variant ID ----
      locuscompare_data <- inner_join(gwas_summary_risklocusN, gtex_locus_gene, by = c("variant_id_b37"))
      if (nrow(locuscompare_data) == 0) next
      remove(gtex_locus_gene)

      ## ---- reformat data for coloc and locus compare ----
      locuscompare_data %<>%
        transmute(
          rsid = SNP,
          logp1 = -log10(P_BOLT_LMM_INF),
          logp2 = -log10(pval_nominal),
          chr = CHR,
          pos = BP,
          beta_gwas = BETA,
          varbeta_gwas = SE^2,
          beta_gtex = slope,
          varbeta_gtex = slope_se^2,
          variant_id_b37,
        )

      ## ---- construct LD matrix for coloc SuSIE ----
      # convert rsid to variant_id_b37
      locuscompare_names <-
        inner_join(
          tibble("rsid" = dimnames(ld_matrix)[[1]]),
          locuscompare_data %>% distinct(rsid, .keep_all = T),
          by = "rsid"
        ) %>%
        select(rsid, variant_id_b37)
      ld_matrix_susie <- ld_matrix[locuscompare_names$rsid, locuscompare_names$rsid]
      dimnames(ld_matrix_susie) <- list(locuscompare_names$variant_id_b37, locuscompare_names$variant_id_b37)

      # make sure LD matrix and locuscompare_data match
      locuscompare_data %<>% filter(variant_id_b37 %in% locuscompare_names$variant_id_b37)

      ## ---- generate coloc data sets ----
      coloc_data_gwas <- list(
        beta = locuscompare_data$beta_gwas,
        varbeta = locuscompare_data$varbeta_gwas,
        snp = locuscompare_data$variant_id_b37,
        position = locuscompare_data$pos,
        type = "quant",
        sdY = sdY_pulsatility,
        LD = ld_matrix_susie
      )
      coloc_data_gtex <- list(
        beta = locuscompare_data$beta_gtex,
        varbeta = locuscompare_data$varbeta_gtex,
        snp = locuscompare_data$variant_id_b37,
        position = locuscompare_data$pos,
        type = "quant",
        sdY = 1,
        LD = ld_matrix_susie
      )
      remove(ld_matrix_susie)

      ## ------------------------------------ RUN COLOC SuSiE ------------------------------------
      # run the SuSiE IBSS algorithm for each dataset, which will identify credible sets of SNPs for GWAS and GTEx
      # susie_gwas <- coloc::runsusie(coloc_data_gwas, coverage = 0.5)
      # susie_gtex <- coloc::runsusie(coloc_data_gtex, coverage = 0.5)
      susie_gwas <- runsusie_silently(coloc_data_gwas, coverage = susie_coverage)
      susie_gtex <- runsusie_silently(coloc_data_gtex, coverage = susie_coverage)
      if (is.null(susie_gwas) || is.null(susie_gtex)) next

      # run coloc to calculate posterior inclusion probabilities
      susie.res <- coloc::coloc.susie(susie_gwas, susie_gtex)
      if (is.null(susie.res$summary)) next

      ## ---- store our results ----
      # do some rearranging, relabeling, renaming to convert our results into an exportable format
      locus_data <-
        tibble(
          fuma_locus_idx = risk_locus_n,
          gtex_tissue = tissue_name,
          gtex_gene = gene_name,
          gtex_ensembl_id = search_gene,
          locus_chr = chr,
          locus_start = min_pos,
          locus_end = max_pos,
          locus_min_gwasP = 10^(-(max(locuscompare_data$logp1))),
          locus_min_gtexP = 10^(-(max(locuscompare_data$logp2))),
        )
      # do some matching to convert the credible sets (which are just lists of the indices of snps) to rsid names
      variant_to_rsid <- locuscompare_names[, c("variant_id_b37", "rsid")] %>% deframe()
      cs_idx_to_rsid <- function(x) paste(variant_to_rsid[names(x)], collapse = ",")
      # compile the coloc susie data into an exportable format
      coloc_susie_data <-
        susie.res$summary %>%
        # rename hit1 from variant_id_b37 format to rsid format
        left_join(locuscompare_names, by = c("hit1" = "variant_id_b37")) %>%
        relocate("rsid", .after = "hit1") %>%
        rename("hit1_rsid" = "rsid", "hit1_variant_id_b37" = "hit1") %>%
        # rename hit2 from variant_id_b37 format to rsid format
        left_join(locuscompare_names, by = c("hit2" = "variant_id_b37")) %>%
        relocate("rsid", .after = "hit2") %>%
        rename("hit2_rsid" = "rsid", "hit2_variant_id_b37" = "hit2") %>%
        # convert the list of snps in each cs to lists of rsids
        left_join(
          summary(susie_gwas)$cs %>%
            mutate(variable = sapply(susie_gwas$sets$cs, cs_idx_to_rsid)) %>%
            rename("rsids" = "variable") %>%
            rename_with(~ paste0("cs1_", str_remove(.x, "^cs_?"))),
          by = c("idx1" = "cs1_")
        ) %>%
        left_join(
          summary(susie_gtex)$cs %>%
            mutate(variable = sapply(susie_gtex$sets$cs, cs_idx_to_rsid)) %>%
            rename("rsids" = "variable") %>%
            rename_with(~ paste0("cs2_", str_remove(.x, "^cs_?"))),
          by = c("idx2" = "cs2_")
        ) %>%
        # prepend column names with coloc.susie_
        rename_with(~ paste0("coloc.susie_", .x))
      # append to our results data frame
      df_coloc.susie_results %<>% bind_rows(bind_cols(locus_data, coloc_susie_data))

      ## ------------------------------------ plot and save data if PP > 0.8 ------------------------------------
      if (any(susie.res$summary$PP.H4.abf > 0.8)) {
        message("         > saving data & plots")

        ## ---- save data ----
        save_path <-
          save_path_root %//%
          str_replace_all(tissue_name, "_", "") %++% "_" %++%
          "chr" %++% chr %++% "_" %++%
          "FumaRiskLocus" %++% formatC(risk_locus_n, width = 2, flag = "0")
        save_name <-
          str_replace_all(tissue_name, "_", "") %++% "_chr" %++% chr %++% "_" %++% gene_name
        dir.create(save_path, showWarnings = F, recursive = T)
        write_rds(susie.res, save_path %//% save_name %++% ".coloc_susie.rds", compress = "gz", compression = 6)
        write_rds(susie_gwas, save_path %//% save_name %++% ".runsusie_gwas.rds", compress = "gz", compression = 6)
        write_rds(susie_gtex, save_path %//% save_name %++% ".runsusie_gtex.rds", compress = "gz", compression = 6)

        ## ---- make sensitivity plots for each hit with PP.H4 > 0.8 ----
        for (i in which(susie.res$summary$PP.H4.abf > 0.8)) {
          pdf(
            file = save_path %//% save_name %++% ".coloc_susie.sensitivity_" %++% formatC(i, width = 2, flag = "0") %++% ".pdf",
            width = 6.5, height = 6.5,
            title = "coloc-SuSiE Sensitivity Analysis"
          )
          suppressMessages(coloc::sensitivity(susie.res, rule = "H4 > 0.8", dataset1 = coloc_data_gwas, dataset2 = coloc_data_gtex, row = i))
          dev.off()
        }

        # identify the snp with the max PPH4
        max_row <- which.max(susie.res$summary$PP.H4.abf)
        snp_max_susie_PPH4 <- variant_to_rsid[susie.res$results$snp[which.max(susie.res$results %>% pull(max_row + 1))]]

        ## ---- make locuscompare plot with each candidate (FUMA) SNP ----
        cand_snps_risklocusN <- fuma_cand_snps %>%
          filter(GenomicLocus == risk_locus_n) %>%
          pull(rsID)
        pb <- txtProgressBar(style = 3, min = 0, max = length(cand_snps_risklocusN))
        for (i in seq_along(cand_snps_risklocusN)) {
          snp_rsID <- cand_snps_risklocusN[i]

          setTxtProgressBar(pb, i)

          is_lead <- any(str_split_fixed(risk_locus$IndSigSNPs, ";", Inf)[1, ] == snp_rsID)
          is_maxPPH4 <- (snp_rsID == snp_max_susie_PPH4)
          targetSNP_in_locus <- any(locuscompare_data$rsid == snp_rsID)

          if (targetSNP_in_locus) {
            save_name <-
              str_replace_all(tissue_name, "_", "") %++% "_chr" %++% chr %++% "_" %++% gene_name %++% "_" %++%
              str_replace_all(snp_rsID, ":", "_") %++% (if (is_lead) "_lead" else "") %++% (if (is_maxPPH4) "_maxPPH4" else "")

            r <- locuscomparer::make_combined_plot(
              locuscompare_data %>% select(rsid, logp1, logp2, chr, pos),
              title1 = "Pulsatility GWAS:",
              title2 = paste0("eQTL ", gene_name, " / ", str_replace_all(tissue_name, "_", " "), ":"),
              ld = tibble(SNP_A = snp_rsID, SNP_B = colnames(ld_matrix), R2 = ld_matrix[match(snp_rsID, colnames(ld_matrix)), ]^2),
              chr = chr,
              snp = snp_rsID,
              combine = FALSE,
              legend = FALSE
            )
            lc_patchwork <- (r$locuscompare + lc_legend) | (r$locuszoom1 / r$locuszoom2) & theme(axis.text = element_text(color = "black"))
            suppressMessages(invisible(ajtools::ggfigsave(
              lc_patchwork,
              name = save_name,
              path = save_path,
              format = c("png", "rds"),
              overwrite = T,
              width = 6.5,
              height = 6.5,
              dpi = 300
            )))
            remove(r, lc_patchwork)
          }
        }
        close(pb)
      }
    }
  }
}

# post-processing; attach some info from the FUMA file about the genomic risk loci & save to disk
df_coloc.susie_results %<>%
  left_join(
    fuma_risk_loci %>%
      select(
        "idx" = GenomicLocus,
        start, end, nSNPs, LeadSNPs
      ) %>%
      rename_with(~ paste0("fuma_locus_", .x)),
    by = "fuma_locus_idx"
  ) %>%
  relocate(starts_with("fuma"), .after = "fuma_locus_idx") %>%
  write_tsv(ajtools::file_datestamp(basename(save_path_root) %++% "_results.txt.gz", path = save_path_root))
```

### Load coloc-SuSiE results

```{r}
df_coloc.susie_results <- vroom(PROJ_ROOT %//% "analyses/coloc_SuSiE_50/coloc_SuSiE_50_results.2021_12_30.txt.gz", show_col_types = F)
```

## Make table of coloc results

First make separate tables for coloc (single) and coloc-SuSiE

```{r, eval=F}
coloc_results_table <- function(df, PP.H4, method_label = NULL) {
  PP.H4 <- as.symbol(PP.H4)
  method_label <- if (!is.null(method_label)) as.symbol(method_label)
  nc <- 5 # number of columns on LHS of table
  df %>%
    mutate(PP.H4 = {{ PP.H4 }}) %>%
    distinct(locus_start, locus_end, gtex_tissue, gtex_gene, gtex_ensembl_id, locus_chr, PP.H4, zscore_cor_top5pct) %>%
    mutate(locus_chr = ordered(locus_chr, levels = c(1:23, "X"), labels = c(1:22, "X", "X"))) %>%
    # group by genes
    group_by(gtex_ensembl_id) %>%
    mutate(max_pph4 = max(PP.H4)) %>%
    ungroup() %>%
    # sort by max pph4
    arrange(-max_pph4, gtex_gene) %>%
    # combine pph4 and zcor
    rowwise() %>%
    mutate(pp_zcor = list(c(PP.H4, zscore_cor_top5pct))) %>%
    ungroup() %>%
    select(-PP.H4, -zscore_cor_top5pct) %>%
    pivot_wider(names_from = gtex_tissue, values_from = pp_zcor, values_fill = list(0, 0)) %>%
    filter(max_pph4 > 0.5) %>%
    transmute(
      "eQTL Gene" = gtex_gene,
      "eQTL Ensembl ID" = xfun::sans_ext(gtex_ensembl_id),
      "Locus \nChr." = locus_chr,
      "Locus\nPosition" = paste0(scibeautify(floor(locus_start)), " ⁠– ", scibeautify(floor(locus_end))),
      "Maximum\nPosterior\nProbability" = max_pph4,
      "Aorta" = Artery_Aorta,
      "Coronary\nArtery" = Artery_Coronary,
      "Tibial\nArtery" = Artery_Tibial,
      "Atrial\nAppendage" = Heart_Atrial_Appendage,
      "Left\nVentricle" = Heart_Left_Ventricle,
      "Lung" = Lung,
      "Cultured\nFibroblasts" = Cells_Cultured_fibroblasts,
    ) %>%
    flextable::flextable() %>%
    flextable::align(j = -1, part = "all", align = "center") %>%
    # alternating row colors
    # flextable::bg(i = ~ seq_len(nrow(data)) %% 2 == 1, bg = "grey97") %>%
    # Merged label over tissues
    flextable::add_header_row(colwidths = c(rep(1, nc), 7), values = c(.$header$col_keys[1:nc], "")) %>%
    flextable::compose(i = 1, j = -c(1:nc), part = "header", value = as_paragraph(
      "Tissue PP", as_sub("H4"), " > 0.5", as_sup("§")
    )) %>%
    flextable::merge_v(j = 1:nc, part = "header") %>%
    flextable::hline(i = 1, j = -c(1:nc), part = "header", border = officer::fp_border(width = 1)) %>%
    # Vertical line between tissues and rest of table
    flextable::vline(j = nc) %>%
    # Footer & Symbols
    flextable::compose(j = "eQTL Gene", part = "header", value = as_paragraph(
      "eQTL Gene", as_sup("†")
    )) %>%
    flextable::add_footer_lines(NA) %>%
    flextable::compose(part = "footer", value = as_paragraph(
      "Chr = chromosome, base pair position in GRCh37 coordinates.",
      "\n",
      "† Gene names are displayed in HUGO gene nomenclature where available, and as Ensembl IDs otherwise.",
      "\n",
      "‡ Posterior probability (PP", as_sub("H4"), ") that there is a shared causal variant in the specified locus for the GWAS and eQTL traits. Only gene-locus pairs with PP", as_sub("H4"), " > 0.5 are displayed. For each gene-locus pair, the maximum PP", as_sub("H4"), " across all tissue types is reported.",
      "\n",
      "§ For each tissue type, a check indicates whether the PP", as_sub("H4"), " for the specified gene-locus pair is greater than 0.5."
    )) %>%
    flextable::line_spacing(part = "footer", space = 1.5) %>%
    flextable::align(j = -1, align = "center", part = "all") %>%
    # PPH4 Format
    flextable::compose(j = "Maximum\nPosterior\nProbability", part = "header", value = as_paragraph(
      "Maximum\nPosterior\nProbability", as_sup("‡")
    )) %>%
    flextable::compose(j = -c(1:(nc - 1)), part = "body", use_dot = T, value = {
      pph4_vec <- sapply(., function(x) x[1] %||% 0)
      zcor_vec <- sapply(., function(x) x[2] %||% NA)
      zcor_sign <- ifelse(pph4_vec > 0.5 & !is.na(zcor_vec) & zcor_vec^2 > 0.8^2, ifelse(zcor_vec > 0, "+", "–"), NA)
      as_paragraph(
        case_when(
          pph4_vec > 0.999 ~ "> 0.999",
          pph4_vec < 0.01 ~ "< 0.01",
          T ~ scibeautify(pph4_vec, 3)
        ),
        " ",
        zcor_sign
      )
    }) %>%
    flextable::bg(j = -c(1:nc), bg = function(x) {
      x <- sapply(x, function(x) x[1] %||% NA)
      ifelse(x < 0.5, NA, {
        rng <- c(0.5, 1)
        x <- (as.numeric(x) - rng[1]) / (rng[2] - rng[1])
        ajtools::col_rmp_blue(x)
      })
    })
}

df_coloc_results %>%
  coloc_results_table("PP.H4") %>%
  ajtools::flextable_save("table_coloc_single_results_v2", path = PROJ_ROOT %//% "figures", overwrite = T)

# df_coloc.susie_results %>%
#   coloc_results_table("coloc.susie_PP.H4.abf") %>%
#   ajtools::flextable_save("table_coloc_SuSiE_results", path = PROJ_ROOT %//% "figures", overwrite = T)
#
# df_coloc_results %>%
#   mutate(method = "single") %>%
#   bind_rows(df_coloc.susie_results %>% mutate(method = "SuSiE", PP.H4 = coloc.susie_PP.H4.abf)) %>%
#   coloc_results_table("PP.H4", "method")
```
